---
title: "Summary RF metrics"
author: "Karin"
date: "2024-05-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, include=FALSE}
library(dplyr)
library(ggplot2)
library(reshape2)
library(tidyr)
library(stringr)
library(MLmetrics)
library(ComplexHeatmap)
library(viridis)
library(scales)
library(wesanderson)
```


# RF10fold, taxonomic levels
## Read in the data
```{r read_files}
base_path = "/Users/emma/Projects/EnvPredict/Results/DifferentTaxonomicLevels"

## Format colnames = sampleid, rownames = parameter

# list and read "Predictions.tsv" files
list_of_prediction_df_files = list.files(base_path, pattern = "Predictions.tsv", full.names = TRUE)
list_of_prediction_dfs = lapply(list_of_prediction_df_files, read.delim)

# list and read "Actual.tsv" files
list_of_true_df_files = list.files(base_path, pattern = "Actual.tsv", full.names = TRUE)
list_of_true_dfs = lapply(list_of_true_df_files, read.delim)

# Function to transpose and format each dataframe
convert_table <- function(df) {
  df <- df %>%
    rownames_to_column(var = "variable") %>%  # Convert row names to a column
    pivot_longer(-variable, names_to = "sample_id", values_to = "value") %>%
    pivot_wider(names_from = variable, values_from = value)
  return(df)
}

# Apply the function to all dataframes in the list
list_of_true_dfs <- lapply(list_of_true_dfs, convert_table)
list_of_prediction_dfs <- lapply(list_of_prediction_dfs, convert_table)

# Check sample ids in predictions vs actual match
list_of_true_dfs[[1]] %>% select(sample_id)
list_of_prediction_dfs[[1]] %>% select(sample_id)

identical(
  list_of_true_dfs[[1]] %>% select(sample_id) %>% pull(),
  list_of_prediction_dfs[[1]] %>% select(sample_id) %>% pull()
)

# list of model names
list_of_true_dfs_name = list.files(base_path, pattern = "Actual.tsv", full.names = FALSE)
model_names = str_remove(list_of_true_dfs_name, "_Actual.tsv")
```

<!---
```{r}
phys_chem_path = '/Users/forskningskarin/Documents/Courses_workshops_etc/EnvGen_hackathon/Results/Summary_physiochemical/physical_chemical_wide_2015_2019_2024-01-17_utf8.txt'

phys_chem = read.delim(phys_chem_path)


phys_chem <- phys_chem %>%
    mutate(institution = case_when(
        grepl("SMHI", dataset_name) ~ "SMHI",
        grepl("UMSC", dataset_name) ~ "UMU",
        grepl("DEEP", dataset_name) ~ "SU",
        TRUE ~ "Other"
    ))

smhi_stations = phys_chem %>%
    filter(institution == "SMHI") %>%
    pull(dataset_name)
```
--->

## Calculate MAPE of predicted and actual values
```{r correlation}
# loop to calculate the correlation for each model
heatmap_df = data.frame()

# set the list of parameters as colnames
for (i in 1:length(list_of_prediction_dfs)){
  
  predictions = list_of_prediction_dfs[[i]]
  true_values = list_of_true_dfs[[i]]
  
  cor_values = c()
  
  # loop to calculate the correlation for each parameter
  for (parameter in list_of_parameters){
    
    filtered_data <- true_values %>%
      filter(!is.na(.data[[parameter]]))
    
    # Get the corresponding predicted values based on the filtered indices
    filtered_predictions <- predictions %>%
      filter(sample_id %in% filtered_data$sample_id)
    
    # Calculate the correlation
    correlation <- MAPE(filtered_predictions[[parameter]], filtered_data[[parameter]])
    cor_values = c(cor_values, correlation)
    
  }
  heatmap_df = rbind(heatmap_df, cor_values)
}

colnames(heatmap_df) = list_of_parameters
rownames(heatmap_df) = model_names
```

## Transform correlation
```{r heatmap}
# convert rownames to a column
heatmap_df$model <- rownames(heatmap_df)

# Melt the dataframe
melted_data <- melt(heatmap_df, id.vars = "model")

# Print the melted dataframe
print(melted_data)
```

## Plot the data
### Heatmap
```{r}
ggplot(melted_data, aes(x = variable, y = model, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Correlation between predicted and true values",
       x = "Parameter",
       y = "Model")
```

### Box plot - all errors
```{r}
melted_data %>% 
  #filter(model == "norm_seqtab_16S_RF10fold") %>%
ggplot(aes(x = variable, y = value, fill = variable)) +
  geom_boxplot() +
  labs(
    x = "Model",
    y = "MAPE Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Make a table of the true and predicted values
```{r}
# Function to add model name to each true dataframe
add_model_to_true_df <- function(true_df, model_name) {
  true_df %>%
    mutate(model = model_name)
}

# Create a large table combining all true values and model names
combined_true_df <- data.frame()

for (i in 1:length(list_of_true_dfs)) {
  true_df <- list_of_true_dfs[[i]]
  model_name <- model_names[i]
  
  # Add the model name to the true_df and bind it to the combined dataframe
  true_df_with_model <- add_model_to_true_df(true_df, model_name)
  combined_true_df <- bind_rows(combined_true_df, true_df_with_model)
}

# View the resulting combined true data frame
head(combined_true_df)

# Melt the dataframe
true_df_long <- combined_true_df %>%
    pivot_longer(cols = -c(sample_id, model), 
                 names_to = "parameter", 
                 values_to = "true_value")

# Print the melted dataframe
head(true_df_long)


# Function to add model name to each predicted dataframe
add_model_to_pred_df <- function(pred_df, model_name) {
  pred_df %>%
    mutate(model = model_name)
}

# Create a large table combining all predicted values and model names
combined_pred_df <- data.frame()

for (i in 1:length(list_of_prediction_dfs)) {
  pred_df <- list_of_prediction_dfs[[i]]
  model_name <- model_names[i]
  
  # Add the model name to the pred_df and bind it to the combined dataframe
  pred_df_with_model <- add_model_to_pred_df(pred_df, model_name)
  combined_pred_df <- bind_rows(combined_pred_df, pred_df_with_model)
}

# View the resulting combined predicted data frame
head(combined_pred_df)

# Melt the dataframe
pred_df_long <- combined_pred_df %>%
    pivot_longer(cols = -c(sample_id, model), 
                 names_to = "parameter", 
                 values_to = "pred_value")

# Print the melted dataframe
head(pred_df_long)


# Merge the true and predicted data frames on the sample_id column
pred_actual_combo <- left_join(true_df_long, pred_df_long, by = c("sample_id", "parameter", "model"))
```
<!---
### Make a scatter of true vs predicted 
```{r}
# Per parameter - single
pred_actual_combo %>% filter(parameter == "Salinity") %>%
ggplot(aes(x = pred_value, y = true_value)) +
  geom_point(alpha = 0.6, color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Add a linear regression line
  facet_wrap(vars(model, parameter), scales = "free") +  # Facet by model and parameter
  labs(x = "True Values", y = "Predicted Values", title = "True vs Predicted") +
  theme_minimal()

# Per parameter - loop
plots_parameter <- map(list_of_parameters, ~ {
  # Filter data for the current parameter
  filtered_data <- pred_actual_combo %>% 
    filter(parameter == .x)
  
  # Generate the plot for this parameter
  ggplot(filtered_data, aes(x = pred_value, y = true_value)) +
    geom_point(alpha = 0.6, color = "blue") +
    geom_smooth(method = "lm", color = "red", se = FALSE) +  # Add a linear regression line
    facet_wrap(vars(model), scales = "free") +  # Facet by model
    labs(x = "True Values", y = "Predicted Values", title = paste("True vs Predicted for", .x)) +
    theme_minimal()
})

# Per model - single
pred_actual_combo %>% filter(model == "norm_clade_counts_16S_2_RF10fold") %>%
ggplot(aes(x = pred_value, y = true_value)) +
  geom_point(alpha = 0.6, color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +  # Add a linear regression line
  facet_wrap(vars(parameter), scales = "free") +  # Facet by model and parameter
  labs(x = "True Values", y = "Predicted Values", title = "True vs Predicted") +
  theme_minimal()

# Per parameter - loop
plots_model <- map(model_names, ~ {
  # Filter data for the current parameter
  filtered_data <- pred_actual_combo %>% 
    filter(model == .x)
  
  # Generate the plot for this parameter
  ggplot(filtered_data, aes(x = pred_value, y = true_value)) +
    geom_point(alpha = 0.6, color = "blue") +
    geom_smooth(method = "lm", color = "red", se = FALSE) +
    facet_wrap(vars(parameter), scales = "free") +
    labs(x = "True Values", y = "Predicted Values", title = paste("True vs Predicted for", .x)) +
    theme_minimal()
})
```
--->

### Scatter plots - true vs predicted
```{r}
# Per model - loop with saving
plots_model <- map(model_names, ~ {
  # Filter data for the current model
  filtered_data <- pred_actual_combo %>% 
    filter(model == .x)
  
  # Generate the plot for this model
  plot <- ggplot(filtered_data, aes(x = pred_value, y = true_value)) +
    geom_point(alpha = 0.6, color = "blue") +
    geom_smooth(method = "lm", color = "red", se = FALSE) +
    facet_wrap(vars(parameter), scales = "free") +
    labs(x = "True Values", y = "Predicted Values", title = paste("True vs Predicted for", .x)) +
    theme_minimal()

  # Save the plot with a dynamic filename (you can adjust the file path)
  ggsave(paste0("Plots/RF10fold/per_model/plot_", .x, ".png"), plot = plot, width = 10, height = 8)
  
  return(plot)
})

# Per parameter - loop with saving
plots_parameter <- map(list_of_parameters, ~ {
  # Filter data for the current parameter
  filtered_data <- pred_actual_combo %>% 
    filter(parameter == .x)
  
  # Generate the plot for this parameter
  plot <- ggplot(filtered_data, aes(x = pred_value, y = true_value)) +
    geom_point(alpha = 0.6, color = "blue") +
    geom_smooth(method = "lm", color = "red", se = FALSE) +  # Add a linear regression line
    facet_wrap(vars(model), scales = "free") +  # Facet by model
    labs(x = "True Values", y = "Predicted Values", title = paste("True vs Predicted for", .x)) +
    theme_minimal()

  # Save the plot with a dynamic filename (you can adjust the file path)
  ggsave(paste0("Plots/RF10fold/per_parameter/plot_", .x, ".png"), plot = plot, width = 10, height = 8)

  return(plot)
})
```

## Add taxonomic levels to data
```{r}
# Pred vs true data
pred_actual_combo <- pred_actual_combo %>%
  mutate(gene = ifelse(str_detect(model, "18S"), "18S", "16S"))

# Add tax levels
pred_actual_tax <- pred_actual_combo %>%
  mutate(tax_level = case_when(
    str_detect(model, "16S_2") ~ "phylum",
    str_detect(model, "16S_3") ~ "class",
    str_detect(model, "16S_4") ~ "order",
    str_detect(model, "16S_5") ~ "family",
    str_detect(model, "16S_6") ~ "genus",
    str_detect(model, "16S_7") ~ "species",
    str_detect(model, "seqtab_16S") ~ "ASV",
    str_detect(model, "18S_2") ~ "supergroup",
    str_detect(model, "18S_3") ~ "division",
    str_detect(model, "18S_4") ~ "subdivision",
    str_detect(model, "18S_5") ~ "class",
    str_detect(model, "18S_6") ~ "order",
    str_detect(model, "18S_7") ~ "family",
    str_detect(model, "18S_8") ~ "genus",
    str_detect(model, "18S_9") ~ "species",
    str_detect(model, "seqtab_18S") ~ "ASV",
    TRUE ~ "other"  # Optional, to cover other cases
  ))

melted_data_tax <- melted_data %>%
  mutate(tax_level = case_when(
    str_detect(model, "16S_2") ~ "phylum",
    str_detect(model, "16S_3") ~ "class",
    str_detect(model, "16S_4") ~ "order",
    str_detect(model, "16S_5") ~ "family",
    str_detect(model, "16S_6") ~ "genus",
    str_detect(model, "16S_7") ~ "species",
    str_detect(model, "seqtab_16S") ~ "ASV",
    str_detect(model, "18S_2") ~ "supergroup",
    str_detect(model, "18S_3") ~ "division",
    str_detect(model, "18S_4") ~ "subdivision",
    str_detect(model, "18S_5") ~ "class",
    str_detect(model, "18S_6") ~ "order",
    str_detect(model, "18S_7") ~ "family",
    str_detect(model, "18S_8") ~ "genus",
    str_detect(model, "18S_9") ~ "species",
    str_detect(model, "seqtab_18S") ~ "ASV",
    TRUE ~ "other"  # Optional, to cover other cases
  ))

# Error data
melted_data_tax <- melted_data_tax %>%
  mutate(amplicon_type = ifelse(str_detect(model, "18S"), "18S", "16S"))

# Define the desired order for each amplicon_type
order_16S <- c("phylum", "class", "order", "family", "genus", "species", "ASV")
order_18S <- c("supergroup", "division", "subdivision", "class", "order", "family", "genus", "species", "ASV")

# Apply the conditional ordering based on amplicon_type
melted_data_tax <- melted_data_tax %>%
  mutate(tax_level = case_when(
    amplicon_type == "16S" ~ factor(tax_level, levels = order_16S),
    amplicon_type == "18S" ~ factor(tax_level, levels = order_18S)))
```

### Box plot - taxonomic levels
```{r}
melted_data_tax %>%
ggplot(aes(x = model, y = value, fill = model)) +
  geom_boxplot() +
  labs(
    x = "Model",
    y = "MAPE Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


melted_data_tax %>% filter(value > 2) %>% distinct(variable)
```

### Box plot - 16S vs 18S
```{r}
# 16S vs 18S
ggplot(melted_data_tax, aes(x = amplicon_type, y = value, fill = amplicon_type)) +
    geom_boxplot(outlier.shape = NA, alpha = 0) +  # Make boxes transparent
    geom_jitter(position = position_jitter(width = 0.2), aes(color = variable, shape = variable), alpha = 0.6, size = 3) +  # Add jittered points
    scale_shape_manual(values = custom_shapes) +  # Use custom shapes for error types
    theme_minimal() +
    # make log y scale
    scale_y_log10() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(title = "16S vs 18S",
         x = "Taxonomic level",
         y = "Error value")

ggsave(plot = last_plot(), filename = "Plots/box_plots/RF10_16S_vs_18S.png")

ggplot(melted_data_tax, aes(x = amplicon_type, y = value, fill = amplicon_type)) +
    geom_boxplot(outlier.shape = NA, alpha = 0) +  # Make boxes transparent
    geom_jitter(position = position_jitter(width = 0.2), aes(color = variable, shape = variable), alpha = 0.6, size = 3) +  # Add jittered points
    scale_shape_manual(values = custom_shapes) +  # Use custom shapes for error types
    theme_minimal() +
    # make log y scale
    scale_y_log10() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(title = "16S vs 18S",
         x = "Taxonomic level",
         y = "Error value") +
    facet_wrap(~ variable, nrow = 1)

ggsave(plot = last_plot(), filename = "Plots/box_plots/RF10_16S_vs_18S_variable.png")
```

### Box plot - 16S, taxonomic levels
```{r}
# 16S taxonomic levels
melted_data_tax %>%
    filter(amplicon_type == "16S") %>%
    ggplot(aes(x = tax_level, y = value, fill = tax_level)) +
    geom_boxplot(outlier.shape = NA, alpha = 0) +  # Make boxes transparent
    geom_jitter(position = position_jitter(width = 0.2), aes(color = variable, shape = variable), alpha = 0.6, size = 3) +  # Add jittered points
    scale_shape_manual(values = custom_shapes) +  # Use custom shapes for error types
    theme_minimal() +
    # make log y scale
    scale_y_log10() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(title = "16S by tax level",
         x = "Taxonomic level",
         y = "Error value")

ggsave(plot = last_plot(), filename = "Plots/box_plots/RF10_16S_tax_level.png")

```

### Box plot - 18S, taxonomic levels
```{r}
# 18S taxonomic levels
melted_data_tax %>%
    filter(amplicon_type == "18S") %>%
    ggplot(aes(x = tax_level, y = value, fill = tax_level)) +
    geom_boxplot(outlier.shape = NA, alpha = 0) + 
    geom_jitter(position = position_jitter(width = 0.2), 
                aes(color = variable, shape = variable), alpha = 0.6, size = 3) +
    scale_shape_manual(values = custom_shapes) +
    theme_minimal() +
    scale_y_log10() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    labs(title = "18S by tax level",
         x = "Taxonomic level",
         y = "Error value")

ggsave(plot = last_plot(), filename = "Plots/box_plots/RF10_18S_tax_level.png")

data_18S <- melted_data_tax %>%
    filter(amplicon_type == "18S")

# Apply the conditional ordering based on amplicon_type
data18S <- data_18S %>%
  mutate(tax_level = case_when(amplicon_type == "18S" ~ factor(tax_level, levels = order_18S)))

data_18S %>%
ggplot(aes(x = tax_level, y = value, fill = tax_level)) +
    geom_boxplot(outlier.shape = NA, alpha = 0) + 
    geom_jitter(position = position_jitter(width = 0.2), 
                aes(color = variable, shape = variable), alpha = 0.6, size = 3) +
    scale_shape_manual(values = custom_shapes) +
    theme_minimal() +
    scale_y_log10() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
    labs(title = "18S by tax level",
         x = "Taxonomic level",
         y = "Error value")

ggsave(plot = last_plot(), filename = "Plots/box_plots/RF10_18S_tax_level.png")

```

# Representations from DeepMicro
## Read in the data
```{r}
base_path = "/Users/emma/Projects/EnvPredict/Results/RepresentationsFromDeepMicro"

# list and read "Predictions.tsv" files
list_of_prediction_df_files = list.files(base_path, pattern = "Predictions.tsv", full.names = TRUE)
list_of_prediction_dfs = lapply(list_of_prediction_df_files, read.delim)

# list and read "Actual.tsv" files
list_of_true_df_files = list.files(base_path, pattern = "Actual.tsv", full.names = TRUE)
list_of_true_dfs = lapply(list_of_true_df_files, read.delim)

# Function to transpose and format each dataframe
convert_table <- function(df) {
  df <- df %>%
    rownames_to_column(var = "variable") %>%  # Convert row names to a column
    pivot_longer(-variable, names_to = "sample_id", values_to = "value") %>%
    pivot_wider(names_from = variable, values_from = value)
  return(df)
}

# Apply the function to all dataframes in the list
list_of_true_dfs <- lapply(list_of_true_dfs, convert_table)
list_of_prediction_dfs <- lapply(list_of_prediction_dfs, convert_table)

# Check sample ids in predictions vs actual match
list_of_true_dfs[[1]] %>% select(sample_id)
list_of_prediction_dfs[[1]] %>% select(sample_id)

identical(
  list_of_true_dfs[[1]] %>% select(sample_id) %>% pull(),
  list_of_prediction_dfs[[1]] %>% select(sample_id) %>% pull()
)

# list of model names
list_of_true_dfs_name = list.files(base_path, pattern = "Actual.tsv", full.names = FALSE)
model_names = str_remove(list_of_true_dfs_name, "_Actual.tsv")
```

## Calculate MAPE of predicted and actual values
```{r correlation}
# loop to calculate the correlation for each model
heatmap_df = data.frame()

# set the list of parameters as colnames
for (i in 1:length(list_of_prediction_dfs)){
  
  predictions = list_of_prediction_dfs[[i]]
  true_values = list_of_true_dfs[[i]]
  
  cor_values = c()
  
  # loop to calculate the correlation for each parameter
  for (parameter in list_of_parameters){
    
    filtered_data <- true_values %>%
      filter(!is.na(.data[[parameter]]))
    
    # Get the corresponding predicted values based on the filtered indices
    filtered_predictions <- predictions %>%
      filter(sample_id %in% filtered_data$sample_id)
    
    # Calculate the correlation
    correlation <- MAPE(filtered_predictions[[parameter]], filtered_data[[parameter]])
    cor_values = c(cor_values, correlation)
    
  }
  heatmap_df = rbind(heatmap_df, cor_values)
}

colnames(heatmap_df) = list_of_parameters
rownames(heatmap_df) = model_names
```

## Transform correlation
```{r heatmap}
# convert rownames to a column
heatmap_df$model <- rownames(heatmap_df)

# Melt the dataframe
melted_data <- melt(heatmap_df, id.vars = "model")

# Print the melted dataframe
print(melted_data)
```

## Plot the data
### Heatmap
```{r}
ggplot(melted_data, aes(x = variable, y = model, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", midpoint = 0.5) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Correlation between predicted and true values",
       x = "Parameter",
       y = "Model")
```

### Box plot - all errors
```{r}
melted_data %>% 
  #filter(model == "norm_seqtab_16S_RF10fold") %>%
ggplot(aes(x = variable, y = value, fill = variable)) +
  geom_boxplot() +
  labs(
    x = "Model",
    y = "MAPE Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Make a table of the true and predicted values
```{r}
# Function to add model name to each true dataframe
add_model_to_true_df <- function(true_df, model_name) {
  true_df %>%
    mutate(model = model_name)
}

# Create a large table combining all true values and model names
combined_true_df <- data.frame()

for (i in 1:length(list_of_true_dfs)) {
  true_df <- list_of_true_dfs[[i]]
  model_name <- model_names[i]
  
  # Add the model name to the true_df and bind it to the combined dataframe
  true_df_with_model <- add_model_to_true_df(true_df, model_name)
  combined_true_df <- bind_rows(combined_true_df, true_df_with_model)
}

# View the resulting combined true data frame
head(combined_true_df)

# Melt the dataframe
true_df_long <- combined_true_df %>%
    pivot_longer(cols = -c(sample_id, model), 
                 names_to = "parameter", 
                 values_to = "true_value")

# Print the melted dataframe
head(true_df_long)


# Function to add model name to each predicted dataframe
add_model_to_pred_df <- function(pred_df, model_name) {
  pred_df %>%
    mutate(model = model_name)
}

# Create a large table combining all predicted values and model names
combined_pred_df <- data.frame()

for (i in 1:length(list_of_prediction_dfs)) {
  pred_df <- list_of_prediction_dfs[[i]]
  model_name <- model_names[i]
  
  # Add the model name to the pred_df and bind it to the combined dataframe
  pred_df_with_model <- add_model_to_pred_df(pred_df, model_name)
  combined_pred_df <- bind_rows(combined_pred_df, pred_df_with_model)
}

# View the resulting combined predicted data frame
head(combined_pred_df)

# Melt the dataframe
pred_df_long <- combined_pred_df %>%
    pivot_longer(cols = -c(sample_id, model), 
                 names_to = "parameter", 
                 values_to = "pred_value")

# Print the melted dataframe
head(pred_df_long)


# Merge the true and predicted data frames on the sample_id column
pred_actual_combo <- left_join(true_df_long, pred_df_long, by = c("sample_id", "parameter", "model"))
```

## Scatter plots - true vs predicted
```{r}
# Per model - loop with saving
plots_model <- map(model_names, ~ {
  # Filter data for the current model
  filtered_data <- pred_actual_combo %>% 
    filter(model == .x)
  
  # Generate the plot for this model
  plot <- ggplot(filtered_data, aes(x = pred_value, y = true_value)) +
    geom_point(alpha = 0.6, color = "blue") +
    geom_smooth(method = "lm", color = "red", se = FALSE) +
    facet_wrap(vars(parameter), scales = "free") +
    labs(x = "True Values", y = "Predicted Values", title = paste("True vs Predicted for", .x)) +
    theme_minimal()

  # Save the plot with a dynamic filename (you can adjust the file path)
  ggsave(paste0("Plots/DeepMicro/per_model/plot_", .x, ".png"), plot = plot, width = 10, height = 8)
  
  return(plot)
})

# Per parameter - loop with saving
plots_parameter <- map(list_of_parameters, ~ {
  # Filter data for the current parameter
  filtered_data <- pred_actual_combo %>% 
    filter(parameter == .x)
  
  # Generate the plot for this parameter
  plot <- ggplot(filtered_data, aes(x = pred_value, y = true_value)) +
    geom_point(alpha = 0.6, color = "blue") +
    geom_smooth(method = "lm", color = "red", se = FALSE) +  # Add a linear regression line
    facet_wrap(vars(model), scales = "free") +  # Facet by model
    labs(x = "True Values", y = "Predicted Values", title = paste("True vs Predicted for", .x)) +
    theme_minimal()

  # Save the plot with a dynamic filename (you can adjust the file path)
  ggsave(paste0("Plots/DeepMicro/per_parameter/plot_", .x, ".png"), plot = plot, width = 10, height = 8)

  return(plot)
})
```

## Add encoding to data
```{r}
# Pred vs true data
pred_actual_combo <- pred_actual_combo %>%
  mutate(amplicon_type = ifelse(str_detect(model, "18S"), "18S", "16S"))

# Add tax levels
pred_actual_combo <- pred_actual_combo %>%
  mutate(encoding = case_when(
    str_detect(model, "^AE_") ~ "AE",
    str_detect(model, "^CAE_") ~ "CAE",
    str_detect(model, "^DAE_") ~ "DAE",
    str_detect(model, "^VAE") ~ "VAE",
    TRUE ~ "other"  # Optional, to cover other cases
  ))

# Error data
melted_data <- melted_data %>%
  mutate(amplicon_type = ifelse(str_detect(model, "18S"), "18S", "16S"))

melted_data <- melted_data %>%
  mutate(encoding = case_when(
    str_detect(model, "^AE_") ~ "AE",
    str_detect(model, "^CAE_") ~ "CAE",
    str_detect(model, "^DAE_") ~ "DAE",
    str_detect(model, "^VAE_") ~ "VAE",
    TRUE ~ "other"  # Optional, to cover other cases
  ))
```

### Box plot - encodings
```{r}
# 16S
melted_data %>%
filter(amplicon_type == "16S") %>%
ggplot(aes(x = model, y = value, fill = encoding)) +
  geom_boxplot() +
  labs(
    x = "Model",
    y = "MAPE Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# 18S
melted_data %>%
filter(amplicon_type == "18S") %>%
ggplot(aes(x = model, y = value, fill = encoding)) +
  geom_boxplot() +
  labs(
    x = "Model",
    y = "MAPE Value"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

melted_data_tax %>% filter(value > 2) %>% distinct(variable)

```

### Box plot - 16S vs 18S
```{r}
# 16S vs 18S
ggplot(melted_data, aes(x = encoding, y = value, fill = encoding)) +
    geom_boxplot(outlier.shape = NA, alpha = 0) +  # Make boxes transparent
    geom_jitter(position = position_jitter(width = 0.2), aes(color = variable, shape = variable), alpha = 0.6, size = 3) +  # Add jittered points
    scale_shape_manual(values = custom_shapes) +  # Use custom shapes for error types
    theme_minimal() +
    # make log y scale
    scale_y_log10() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(title = "16S vs 18S",
         x = "Taxonomic level",
         y = "Error value") +
  facet_wrap(~ amplicon_type, nrow = 1)

ggsave(plot = last_plot(), filename = "Plots/DeepMicro/box_plots/Encoding_16S_vs_18S.png")

ggplot(melted_data, aes(x = encoding, y = value, fill = encoding)) +
    geom_boxplot(outlier.shape = NA, alpha = 0) +  # Make boxes transparent
    geom_jitter(position = position_jitter(width = 0.2), aes(color = variable, shape = variable), alpha = 0.6, size = 3) +  # Add jittered points
    scale_shape_manual(values = custom_shapes) +  # Use custom shapes for error types
    theme_minimal() +
    # make log y scale
    scale_y_log10() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(title = "Encoding vs variable",
         x = "Taxonomic level",
         y = "Error value") +
    facet_wrap(~ variable, nrow = 1)

ggsave(plot = last_plot(), filename = "Plots/DeepMicro/box_plots/Encoding_vs_variable.png")
```

### Box plot - 16S, 18S, encodings
```{r}
# 16S taxonomic levels
melted_data %>%
    filter(amplicon_type == "16S") %>%
    ggplot(aes(x = encoding, y = value, fill = encoding)) +
    geom_boxplot(outlier.shape = NA, alpha = 0) +  # Make boxes transparent
    geom_jitter(position = position_jitter(width = 0.2), aes(color = variable, shape = variable), alpha = 0.6, size = 3) +  # Add jittered points
    scale_shape_manual(values = custom_shapes) +  # Use custom shapes for error types
    theme_minimal() +
    # make log y scale
    scale_y_log10() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(title = "16S by tax level",
         x = "Taxonomic level",
         y = "Error value")

ggsave(plot = last_plot(), filename = "Plots/DeepMicro/box_plots/Encoding_16S_variable.png")

# 18S taxonomic levels
melted_data %>%
    filter(amplicon_type == "18S") %>%
    ggplot(aes(x = encoding, y = value, fill = encoding)) +
    geom_boxplot(outlier.shape = NA, alpha = 0) +  # Make boxes transparent
    geom_jitter(position = position_jitter(width = 0.2), aes(color = variable, shape = variable), alpha = 0.6, size = 3) +  # Add jittered points
    scale_shape_manual(values = custom_shapes) +  # Use custom shapes for error types
    theme_minimal() +
    # make log y scale
    scale_y_log10() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(title = "18S by tax level",
         x = "Taxonomic level",
         y = "Error value")

ggsave(plot = last_plot(), filename = "Plots/DeepMicro/box_plots/Encoding_18S_variable.png")
```

# Compare RF and DeepMicro
```{r}
RF10F <- melted_data_tax %>% filter(str_detect(model, "seqtab"))
RF10F <- RF10F %>% 
  mutate(encoding = "RF10fold") %>%
  select(-c(tax_level))

compare_methods <- bind_rows(melted_data, RF10F)
```

## Plot
```{r}
compare_methods %>%
    filter(amplicon_type == "16S") %>%
    ggplot(aes(x = encoding, y = value, fill = encoding)) +
    geom_boxplot(outlier.shape = NA, alpha = 0) +  # Make boxes transparent
    geom_jitter(position = position_jitter(width = 0.2), aes(color = variable, shape = variable), alpha = 0.6, size = 3) +  # Add jittered points
    scale_shape_manual(values = custom_shapes) +  # Use custom shapes for error types
    theme_minimal() +
    # make log y scale
    scale_y_log10() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(title = "16S",
         x = "Model/encoding",
         y = "Error value")

compare_methods %>%
    ggplot(aes(x = encoding, y = value, fill = encoding)) +
    geom_boxplot(outlier.shape = NA, alpha = 0) +  # Make boxes transparent
    geom_jitter(position = position_jitter(width = 0.2), aes(color = variable, shape = variable), alpha = 0.6, size = 3) +  # Add jittered points
    scale_shape_manual(values = custom_shapes) +  # Use custom shapes for error types
    theme_minimal() +
    # make log y scale
    scale_y_log10() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(title = "Compare",
         x = "Model/encoding",
         y = "Error value") +
  facet_wrap(~ amplicon_type, nrow = 1)

ggsave(plot = last_plot(), filename = "Plots/Compare/box_plots/plot_RF_vs_DeepMicro.png")
```



<!---
```{r}
# 18S taxonomic levels
melted_data_tax %>%
    filter(amplicon_type == "18S") %>%
    ggplot(aes(x = tax_level, y = value, fill = tax_level)) +
    geom_boxplot(outlier.shape = NA, alpha = 0) +  # Make boxes transparent
    geom_jitter(position = position_jitter(width = 0.2), aes(color = variable, shape = variable), alpha = 0.6, size = 3) +  # Add jittered points
    scale_shape_manual(values = custom_shapes) +  # Use custom shapes for error types
    theme_minimal() +
    # make log y scale
    scale_y_log10() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
    labs(title = "18S by tax level",
         x = "Taxonomic level",
         y = "Error value")
```


```{r}

# Reshape the data frame from wide to long format
long_model_performance <- heatmap_anno_cols %>%
  pivot_longer(cols = c(1:(ncol(heatmap_anno_cols) - 5)),
              names_to = "error_type", 
               values_to = "error_value") %>%
  mutate(tax_level = factor(tax_level, levels = c("ASV", "Species", "Genus", "Family", "Order", "Class")))
  
# 16S = 2 class
# 18S = 8 speces
```

```{r}
# create a box plot based on the taxonomic level (no nitrogen)
filtered_data_no_nitrogen <- long_model_performance %>%
  filter(error_type != 'NO3_NO2')


# Define custom shapes for the error types
custom_shapes <- c(16, 17, 18, 19, 15, 8, 3, 4, 7, 9, 10, 12, 13, 14, 20, 21, 1, 2, 25, 5)



# Filter the data to exclude 'NO3_NO2'
#filtered_data_no_nitrogen <- long_model_performance %>% filter(error_type != 'NO3_NO2')

filtered_data_no_nitrogen = long_model_performance
# Create the plot with transparent boxes and jittered points, using custom shapes
ggplot(filtered_data_no_nitrogen, aes(x = amplicon_type, y = error_value, fill = amplicon_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0) +  # Make boxes transparent
  geom_jitter(position = position_jitter(width = 0.2), aes(color = error_type, shape = error_type), alpha = 0.6, size = 3) +  # Add jittered points
  scale_shape_manual(values = custom_shapes) +  # Use custom shapes for error types
  theme_minimal() +
  # make log y scale
  scale_y_log10() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Model performance based on taxonomic level",
       x = "Taxonomic level",
       y = "Error value") +
  facet_wrap(~ tax_level, nrow = 1)

```
--->


```{r}
# Create the plot with transparent boxes and jittered points, using custom shapes
ggplot(filtered_data_no_nitrogen, aes(x = amplicon_type, y = error_value, fill = amplicon_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0) +  # Make boxes transparent
  geom_jitter(position = position_jitter(width = 0.2), aes(color = error_type, shape = error_type), alpha = 0.6, size = 3) +  # Add jittered points
  scale_shape_manual(values = custom_shapes) +  # Use custom shapes for error types
  theme_minimal() +
  # make log y scale
  scale_y_log10() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Model performance based on type of encoding",
       x = "Taxonomic level",
       y = "Error value") +
  facet_wrap(~ encoding, nrow = 1)
```

```{r}
# Create the plot with transparent boxes and jittered points, using custom shapes
ggplot(filtered_data_no_nitrogen, aes(x = amplicon_type, y = error_value, fill = amplicon_type)) +
  geom_boxplot(outlier.shape = NA, alpha = 0) +  # Make boxes transparent
  geom_jitter(position = position_jitter(width = 0.2), aes(color = error_type, shape = error_type), alpha = 0.6, size = 3) +  # Add jittered points
  scale_shape_manual(values = custom_shapes) +  # Use custom shapes for error types
  theme_minimal() +
  scale_y_log10() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  labs(title = "Model performance based on type of amplicon",
       x = "Taxonomic level",
       y = "Error value")

```


```{r}
# compare the 16S and 
test_df = long_model_performance %>%
  group_by(amplicon_type) %>%
  summarise(mean_error = mean(error_value), sd_error = sd(error_value))

perc = test_df$mean_error[test_df$amplicon_type == '18S'] / test_df$mean_error[test_df$amplicon_type == '16S']
print(paste('The 18S models have mean MAPE values that are', round(perc, 2), 'times higher than the 16S models'))
```

```{r}
best_model_name = 'RF_withCV_mtryopt_ASV_16S_Abiotic'

actual_values_file_name = paste0(best_model_name, '_Actual_values.tsv')
predictions_file_name = paste0(best_model_name, '_Predictions.tsv')

actual_values = read.delim(paste0(base_path, '/',actual_values_file_name))
predictions = read.delim(paste0(base_path, '/',predictions_file_name))
```


```{r}
# merge the data
predictions_and_actual_data = merge(actual_values, predictions, by = 'sample_id')
```

```{r}
# plot the salinity and predicted salinity
ggplot() +
  geom_point(data = predictions_and_actual_data, aes(x = Salinity, y = Predicted_Salinity), color = 'blue') +
  geom_abline(intercept = 0, slope = 1, color = 'red') +
  labs(title = 'Actual vs Predicted Salinity',
       x = 'Actual Salinity',
       y = 'Predicted Salinity')


# plot the salinity and predicted salinity
ggplot() +
  geom_point(data = predictions_and_actual_data, aes(x = NO3_NO2, y = Predicted_NO3_NO2), color = 'blue') +
  geom_abline(intercept = 0, slope = 1, color = 'red') +
  labs(title = 'Actual vs Predicted NO3 and NO2',
       x = 'Actual NO3_NO2',
       y = 'Predicted NO3_NO2')

# plot the pH and predicted pH
ggplot() +
  geom_point(data = predictions_and_actual_data, aes(x = pH, y = Predicted_pH), color = 'blue') +
  geom_abline(intercept = 0, slope = 1, color = 'red') +
  labs(title = 'Actual vs Predicted pH',
       x = 'Actual pH',
       y = 'Predicted pH')


ggplot() +
  geom_point(data = predictions_and_actual_data, aes(x = Phosphate, y = Predicted_Phosphate), color = 'blue') +
  geom_abline(intercept = 0, slope = 1, color = 'red') +
  labs(title = 'Actual vs Predicted Phosphate',
       x = 'Actual Phosphate',
       y = 'Predicted Phosphate')
```





```{r}
# look at biotic factors
base_path = "/Users/forskningskarin/Documents/Courses_workshops_etc/EnvGen_hackathon/Results/Summary17jan2025/Biotic"

# list and read "Predictions.tsv" files
list_of_prediction_dfs_name = list.files(base_path, pattern = "Predictions.tsv", full.names = TRUE)
list_of_prediction_dfs = lapply(list_of_prediction_dfs_name, read.delim)

# rename the columns
for (i in 1:length(list_of_prediction_dfs)){
  colnames(list_of_prediction_dfs[[i]]) <- str_remove(colnames(list_of_prediction_dfs[[i]]), "Predicted_")
}

list_of_parameters = colnames(list_of_prediction_dfs[[i]])
list_of_parameters = list_of_parameters[!list_of_parameters %in% c("sample_id")]

# list and read "Actual_values.tsv" files
list_of_true_dfs_name = list.files(base_path, pattern = "Actual_values.tsv", full.names = TRUE)
# read the files
list_of_true_dfs = lapply(list_of_true_dfs_name, read.delim)

# list of model names
list_of_true_dfs_name = list.files(base_path, pattern = "Actual_values.tsv", full.names = FALSE)
model_names = str_remove(list_of_true_dfs_name, "_Actual_values.tsv")
```


```{r}
heatmap_df <- data.frame()

# Loop to calculate the correlation for each model
for (i in seq_along(list_of_prediction_dfs)) {
  
  predictions <- list_of_prediction_dfs[[i]]
  true_values <- list_of_true_dfs[[i]]
  
  # Ensure shared columns between predictions and true_values
  common_parameters <- intersect(colnames(predictions), colnames(true_values))
  
  # if sample id is a parameter, remove it
  common_parameters <- common_parameters[common_parameters != 'sample_id']
  cor_values <- numeric(length(common_parameters))
  
  for (j in seq_along(common_parameters)) {
    parameter <- common_parameters[j]
    
    # Filter true values and ensure no NA
    filtered_data <- true_values %>%
      filter(!is.na(.data[[parameter]]))
    
    # Match predictions by sample_id
    filtered_predictions <- predictions %>%
      filter(sample_id %in% filtered_data$sample_id) %>%
      arrange(match(sample_id, filtered_data$sample_id))
    
    # Calculate correlation (replace MAPE with your actual function)
    correlation <- MAPE(filtered_predictions[[parameter]], filtered_data[[parameter]])
    cor_values[j] <- correlation
  }
  
  # Add correlations to heatmap_df
  heatmap_df <- rbind(heatmap_df, cor_values)
}

# Set column and row names
colnames(heatmap_df) <- common_parameters
rownames(heatmap_df) <- model_names
```



