---
title: "Evaluation of cross-dataset predicitons, linear regression"
author: "Krzysztof Jurdzinski"
date: "2025-07-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


```{r packages, include=FALSE}
library(dplyr)
library(ggplot2)
library(reshape2)
library(tidyr)
library(stringr)
library(MLmetrics)
library(ComplexHeatmap)
library(viridis)
library(scales)
library(wesanderson)
```


## Define and load infiles, define outfiles

```{r read_files}
base_path = "../output/predict_2015_2017/"

# list and read "Predictions.tsv" files
list_of_prediction_dfs_name = list.files(base_path, pattern = "Predictions.tsv", full.names = TRUE)
list_of_prediction_dfs = lapply(list_of_prediction_dfs_name, read.delim)

# rename the columns
for (i in 1:length(list_of_prediction_dfs)){
  colnames(list_of_prediction_dfs[[i]]) <- str_remove(colnames(list_of_prediction_dfs[[i]]), "Predicted_")
}

list_of_parameters = colnames(list_of_prediction_dfs[[i]])
list_of_parameters = list_of_parameters[!list_of_parameters %in% c("sample_id")]

# list and read "Actual_values.tsv" files
list_of_true_dfs_name = list.files(base_path, pattern = "Actual.tsv", full.names = TRUE)
# read the files
list_of_true_dfs = lapply(list_of_true_dfs_name, read.delim)

# list of model names
list_of_true_dfs_name = list.files(base_path, pattern = "Actual.tsv", full.names = FALSE)
model_names = str_remove(list_of_true_dfs_name, "_Actual.tsv")

model_names_base = str_remove(model_names, "_2015_2017")

plot_folder = '../output/interannual_comparison_plots/'
if (!dir.exists(plot_folder)){
  dir.create(plot_folder)
}
```

Compare actual phys_chem values 2015-2017 predicted 

```{r}
predictions = list_of_prediction_dfs[[which(model_names == 'norm_asv_counts_16S_phys_chem_2015_2017_different_years')]]
true_values = list_of_true_dfs[[which(model_names == 'norm_asv_counts_16S_phys_chem_2015_2017_different_years')]]

list_of_parameters = rownames(predictions)

identical(colnames(predictions), colnames(true_values))

cor_values = c()

plot_subfolder = paste0(plot_folder, "different_dataset/")
if (!dir.exists(plot_subfolder)){
  dir.create(plot_subfolder)
}

for(parameter in list_of_parameters){
    
  
  data = data.frame(true_values = unlist(true_values[parameter,]), predictions = unlist(predictions[parameter,]))
  data = data[complete.cases(data),]
  
  if(nrow(data) == 0){
    next
  }
  
  # correlation = cor(data$predictions, data$true_values)
  correlation = round(summary(lm(data$predictions ~ data$true_values))$r.squared, 3)
  
  cor_values = c(cor_values, correlation)
  names(cor_values)[length(cor_values)] = parameter
}

sort(cor_values)

```

Get the correlations for the 16S ASV based CV predictions

```{r}
predictions = list_of_prediction_dfs[[which(model_names == 'norm_asv_counts_16S_phys_chem_2015_2017_both_years')]]
true_values = list_of_true_dfs[[which(model_names == 'norm_asv_counts_16S_phys_chem_2015_2017_both_years')]]

list_of_parameters = rownames(predictions)

true_values = true_values[,colnames(predictions)]

identical(colnames(predictions), colnames(true_values))

cor_values_both_datasets = c()

for(parameter in list_of_parameters){
  
  
  data = data.frame(true_values = unlist(true_values[parameter,]), predictions = unlist(predictions[parameter,]))
  data = data[complete.cases(data),]
  
  if(nrow(data) == 0){
    next
  }
  
  correlation = round(summary(lm(data$predictions ~ data$true_values))$r.squared, 3)
  
  cor_values_both_datasets = c(cor_values_both_datasets, correlation)
  names(cor_values_both_datasets)[length(cor_values_both_datasets)] = parameter
}

sort(cor_values_both_datasets)

```
Get the correlations for the 16S ASV based CV predictions: 2019-2020 training and testing data

```{r}
predictions = list_of_prediction_dfs[[which(model_names == 'norm_asv_counts_16S_phys_chem_2015_2017_same_year')]]
true_values = list_of_true_dfs[[which(model_names == 'norm_asv_counts_16S_phys_chem_2015_2017_same_year')]]

list_of_parameters = rownames(predictions)

true_values = true_values[,colnames(predictions)]

identical(colnames(predictions), colnames(true_values))

cor_values_same_dataset = c()


for(parameter in list_of_parameters){
  
  
  data = data.frame(predictions = unlist(predictions[parameter,]), true_values = unlist(true_values[parameter,]))
  data = data[complete.cases(data),]
  
  if(nrow(data) == 0){
    next
  }
  
  correlation = round(summary(lm(data$predictions ~ data$true_values))$r.squared, 3)
  
  cor_values_same_dataset = c(cor_values_same_dataset, correlation)
  names(cor_values_same_dataset)[length(cor_values_same_dataset)] = parameter
}

sort(cor_values_same_dataset)

```


## Compare the values between the different_dataset and the cv prediction

```{r}
cor_df = data.frame(parameter = names(cor_values), both_datasets = cor_values_both_datasets, same_dataset = cor_values_same_dataset, different_dataset = cor_values)

ix = which(cor_df$parameter == 'Longitude')

cor_df = cor_df[ix:nrow(cor_df),]

cor_df = cor_df[complete.cases(cor_df),]

pdf(paste0(plot_folder, "phys_chem_interannual_comparison.pdf"))

layout(matrix(c(1,2,1,3), 2, 2, byrow = T))

group_colors <- c('#edf8b1', '#7fcdbb', '#2c7fb8')
names(group_colors) <- c('Different dataset', 'Same dataset', 'Both datasets')

par(mar = c(4,10,2,2), xpd = T)
barplot(as.matrix(t(cor_df[,c(2,3,4)])), beside = T, horiz = T, las = 1, cex.names = 0.8,
        ## Make it list all the names
        names.arg = cor_df$parameter,
        xlab = expression(paste("Coefficient of determination (", R^2, ")")),
        xlim = c(0,1), col = rev(group_colors)
        ## Move legend
         # legend = T, args.legend = list(x = "topleft", inset = c(-1,0))
        )
legend("topleft", inset = c(-0.4, 0),
       legend = c('Different dataset', "Same dataset", "Both dataset"),
       fill = group_colors,
       title = "Model trained on:")
par(mar = c(6,4,2,2))
boxplot(cor_df$different_dataset, cor_df$same_dataset, cor_df$both_datasets,  names = c('Different dataset', "Same dataset", "Both datasets"), las = 2, col = group_colors, ylim = c(0,1),
        ylab = expression(paste("Coefficient of determination (", R^2, ")")))
par(mar = c(4,4,2,2))
plot(cor_df$different_dataset, cor_df$same_dataset, xlab = 'Different dataset', ylab = 'Same datasets', main = 'Different dataset vs both datasets correlation coefficients',
     col = '#378a76ff', bg = group_colors[1], pch = 21,
     ## Add an identity line
     xlim = c(0,1), ylim = c(0,1))
lines(c(0,1), c(0,1))
dev.off()

wilcox.test(cor_df$different_dataset, cor_df$same_dataset, paired = T)
median(cor_df$same_dataset) - median(cor_df$different_dataset)

wilcox.test(cor_df$different_dataset, cor_df$both_datasets, paired = T)
median(cor_df$both_datasets) - median(cor_df$different_dataset)

wilcox.test(cor_df$both_datasets, cor_df$same_dataset, paired = T)
median(cor_df$both_datasets) - median(cor_df$same_dataset)

## Get only rows in which max R^2 is larger than 0.8
ix = which(apply(cor_df[,c(3,4)], 1, max) > 0.8)

wilcox.test(cor_df$same_dataset[ix], cor_df$different_dataset[ix], paired = T)
median(cor_df$same_dataset[ix]) - median(cor_df$different_dataset[ix])

```
Check for which parameters using both datasets is not the best strategy

```{r}
ix = which(cor_df$both_datasets < cor_df$same_dataset & cor_df$both_datasets < cor_df$different_dataset)

both_worst = cor_df[ix,]

ix = which(cor_df$both_datasets < cor_df$same_dataset)

both_worse_than_same = cor_df[ix,]
```

