---
title: "Evaluation of interannual predicitions"
author: "Krzysztof"
date: "2025-03-28"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown


```{r packages, include=FALSE}
library(dplyr)
library(ggplot2)
library(reshape2)
library(tidyr)
library(stringr)
library(MLmetrics)
library(ComplexHeatmap)
library(viridis)
library(scales)
library(wesanderson)
```


```{r read_files}
base_path = "../output/predict_2015_2017/"

# list and read "Predictions.tsv" files
list_of_prediction_dfs_name = list.files(base_path, pattern = "Predictions.tsv", full.names = TRUE)
list_of_prediction_dfs = lapply(list_of_prediction_dfs_name, read.delim)

# rename the columns
for (i in 1:length(list_of_prediction_dfs)){
  colnames(list_of_prediction_dfs[[i]]) <- str_remove(colnames(list_of_prediction_dfs[[i]]), "Predicted_")
}

list_of_parameters = colnames(list_of_prediction_dfs[[i]])
list_of_parameters = list_of_parameters[!list_of_parameters %in% c("sample_id")]

# list and read "Actual_values.tsv" files
list_of_true_dfs_name = list.files(base_path, pattern = "Actual.tsv", full.names = TRUE)
# read the files
list_of_true_dfs = lapply(list_of_true_dfs_name, read.delim)

# list of model names
list_of_true_dfs_name = list.files(base_path, pattern = "Actual.tsv", full.names = FALSE)
model_names = str_remove(list_of_true_dfs_name, "_Actual.tsv")

model_names_base = str_remove(model_names, "_2015_2017")

plot_folder = '../output/interannual_comparison_plots/'
if (!dir.exists(plot_folder)){
  dir.create(plot_folder)
}



```


Load the 16S ASV based predicitons and actual data: Training on the whole dataset

```{r}
base_path_phys_chem = "../output/DifferentTaxonomicLevels/"

phys_chem_cv_predicitons = read.delim(paste0(base_path_phys_chem, "norm_seqtab_16S_RF10fold_Predictions.tsv"))
phys_chem_cv_actual = read.delim(paste0(base_path_phys_chem, "norm_seqtab_16S_RF10fold_Actual.tsv"))

base_path_phyt_plan = "../output/phytoplankton_predicted/"

phyt_plan_genus_cv_predicitons = read.delim(paste0(base_path_phyt_plan, "norm_asv_counts_18S_phyt_plan_genus_RF10fold_Predictions.tsv"))
phyt_plan_genus_cv_actual = read.delim(paste0(base_path_phyt_plan, "norm_asv_counts_18S_phyt_plan_genus_RF10fold_Actual.tsv"))

base_path_zoo_plan = "../output/zooplankton_predicted/"

zoo_plan_genus_cv_predicitons = read.delim(paste0(base_path_zoo_plan, "norm_asv_counts_16S_zoo_plan_genus_RF10fold_Predictions.tsv"))
zoo_plan_genus_cv_actual = read.delim(paste0(base_path_zoo_plan, "norm_asv_counts_16S_zoo_plan_genus_RF10fold_Actual.tsv"))

```

Load the 16S ASV based predicitons and actual data: 2019-2020 training and prediction

```{r}
base_path_2019_2020 = "../output/only_2019_2020/"

phys_chem_cv_predicitons_2019 = read.delim(paste0(base_path_2019_2020, "norm_asv_counts_16S_phys_chem_2019_2020_Predictions.tsv"))
phys_chem_cv_actual_2019 = read.delim(paste0(base_path_2019_2020, "norm_asv_counts_16S_phys_chem_2019_2020_Actual.tsv"))

phyt_plan_genus_cv_predicitons_2019 = read.delim(paste0(base_path_2019_2020, "norm_asv_counts_18S_phyt_plan_genus_2019_2020_Predictions.tsv"))
phyt_plan_genus_cv_actual_2019 = read.delim(paste0(base_path_2019_2020, "norm_asv_counts_18S_phyt_plan_genus_2019_2020_Actual.tsv"))

zoo_plan_genus_cv_predicitons_2019 = read.delim(paste0(base_path_2019_2020, "norm_asv_counts_16S_zoo_plan_genus_2019_2020_Predictions.tsv"))
zoo_plan_genus_cv_actual_2019 = read.delim(paste0(base_path_2019_2020, "norm_asv_counts_16S_zoo_plan_genus_2019_2020_Actual.tsv"))


```



```{r}
phys_chem_path = '../env_data/combined/physical_chemical_wide_2015_2019_2024-01-17_utf8.txt'

phys_chem = read.delim(phys_chem_path)


phys_chem <- phys_chem %>%
    mutate(institution = case_when(
        grepl("SMHI", dataset_name) ~ "SMHI",
        grepl("UMSC", dataset_name) ~ "UMU",
        grepl("DEEP", dataset_name) ~ "SU",
        TRUE ~ "Other"
    ))

smhi_stations = phys_chem %>%
    filter(institution == "SMHI") %>%
    pull(dataset_name)
```


Compare acrual phys_chem values

```{r}
predictions = list_of_prediction_dfs[[which(model_names == 'norm_asv_counts_16S_phys_chem_2015_2017')]]
true_values = list_of_true_dfs[[which(model_names == 'norm_asv_counts_16S_phys_chem_2015_2017')]]

list_of_parameters = rownames(predictions)

identical(colnames(predictions), colnames(true_values))

cor_values = c()

for(parameter in list_of_parameters){
    
  
  data = data.frame(predictions = unlist(true_values[parameter,]), true_values = unlist(predictions[parameter,]))
  data = data[complete.cases(data),]
  
  correlation = cor(data$predictions, data$true_values)
  
  cor_values = c(cor_values, correlation)
  names(cor_values)[length(cor_values)] = parameter
}

cor_values

```
Visualize the correlation

```{r}
par(mfcol = c(1,1), mar = c(2,10,2,20), xpd = T)
barplot(cor_values, beside = T, horiz = T, las = 1, cex.names = 0.8,
        ## Make it list all the names
        names.arg = names(cor_values),
        ##Make x-axis scale from 0 to 1
        xlim = c(0,1)
        )
# barplot(t(cor_matr), beside = T, horiz = T, las = 1, legend = T, names.arg = sample_labels, cex.names = 0.8, args.legend = list(x = "bottom", inset = c(1,-0.25)))

```

Get the correlations for the 16S ASV based CV predictions

```{r}
cor_values_cv = c()

identical(rownames(phys_chem_cv_predicitons), rownames(predictions))

for(parameter in list_of_parameters){
    
  
  data = data.frame(predictions = unlist(phys_chem_cv_actual[parameter,]), true_values = unlist(phys_chem_cv_predicitons[parameter,]))
  data = data[complete.cases(data),]
  
  correlation = cor(data$predictions, data$true_values)
  
  cor_values_cv = c(cor_values_cv, correlation)
  names(cor_values_cv)[length(cor_values_cv)] = parameter
}

```
Get the correlations for the 16S ASV based CV predictions: 2019-2020 training and testing data

```{r}
cor_values_cv_2019 = c()

identical(rownames(phys_chem_cv_predicitons_2019), rownames(predictions))

for(parameter in list_of_parameters){
    
  
  data = data.frame(predictions = unlist(phys_chem_cv_actual_2019[parameter,]), true_values = unlist(phys_chem_cv_predicitons_2019[parameter,]))
  data = data[complete.cases(data),]
  
  correlation = cor(data$predictions, data$true_values)
  
  cor_values_cv_2019 = c(cor_values_cv_2019, correlation)
  names(cor_values_cv_2019)[length(cor_values_cv_2019)] = parameter
}

```


## Compare the values between the interannual and the cv prediction

```{r}
cor_df = data.frame(parameter = names(cor_values), cv = cor_values_cv, cv_2019 = cor_values_cv_2019, interannual = cor_values)

ix = which(cor_df$parameter == 'Longitude')

cor_df = cor_df[ix:nrow(cor_df),]

cor_df = cor_df[complete.cases(cor_df),]

pdf(paste0(plot_folder, "phys_chem_cv_interannual_comparison.pdf"))

layout(matrix(c(1,2,1,3), 2, 2, byrow = T))

par(mar = c(2,10,2,2), xpd = T)
barplot(as.matrix(t(cor_df[,c(2,3,4)])), beside = T, horiz = T, las = 1, cex.names = 0.8,
        ## Make it list all the names
        names.arg = cor_df$parameter,
        ##Make x-axis scale from 0 to 1
        xlim = c(0,1),
        ## Move legend
         legend = T, args.legend = list(x = "topleft", inset = c(-1,0))
        )
par(mar = c(6,4,2,2))
boxplot(cor_df$interannual, cor_df$cv_2019, cor_df$cv,  names = c('CV', 'CV 2019-2020', 'Interannual'), las = 2)
par(mar = c(4,4,2,2))
plot(cor_df$interannual, cor_df$cv_2019, xlab = 'Interannual', ylab = 'CV 2019-2020', main = 'Interannual vs CV 2019-2020 correlation',
     ## Add an identity line
     xlim = c(0,1), ylim = c(0,1))
lines(c(0,1), c(0,1))
dev.off()


wilcox.test(cor_df$interannual, cor_df$cv, paired = T)
median(cor_df$cv) - median(cor_df$interannual)

wilcox.test(cor_df$interannual, cor_df$cv_2019, paired = T)
median(cor_df$cv_2019) - median(cor_df$interannual)

wilcox.test(cor_df$cv_2019, cor_df$cv, paired = T)
median(cor_df$cv_2019) - median(cor_df$cv)

```



