---
title: "Plot_Results"
author: "Emma"
date: "2024-03-28"
output: html_document
---

# Packages
```{r packages, include=FALSE}
library(tidyverse)
library(purrr)
library(stringr)
library(MLmetrics)
library(scales)
library(ggpubr)
library(cowplot)
library(tools)
library(rstatix)
```
# Predictions
```{r}
# Random forest - all taxonomic levels
RF_dir <- "/Users/emma/Projects/EnvPredict/Predictions/DifferentTaxonomicLevels"

# TabPNF - ASV level only
PNF_dir <- "/Users/emma/Projects/EnvPredict/Predictions/TabPNF/DifferentTaxonomicLevels"

# Representations from deep micro
Deep_dir <- "/Users/emma/Projects/EnvPredict/Predictions/RepresentationsFromDeepMicro_prefilt_features"

# XGBoost
XGB_dir <- "/Users/emma/Projects/EnvPredict/Predictions/XGBoost_ASVs"
```

# Parameters predicted
```{r}
# list parameters to plot
list_of_parameters <- c("Longitude", "Latitude", "Salinity", "Temperature", "pH", "Alkalinity", "Secchi_depth", "SiO3", "N_tot", "DIN", "NH4", "NO2", "NO3", "NO3_NO2", "P_tot", "Phosphate", "DOC", "Humus", "Chl")

# Define custom shapes
custom_shapes <- c(16, 17, 18, 19, 15, 8, 3, 4, 7, 9, 10, 12, 13, 14, 20, 21, 22, 0, 1)

physical_parameters <- c(
  "Longitude", "Latitude", "Salinity", "Temperature", "pH",
  "Alkalinity", "Secchi_depth"
)

nutrient_parameters <- c(
  "SiO3", "N_tot", "DIN", "NH4", "NO2", "NO3", "NO3_NO2",
  "P_tot", "Phosphate", "DOC", "Humus", "Chl"
)

gene_labels <- c(
  "16S" = "16S rRNA",
  "18S" = "18S rRNA"
)
```

# Read in data
## Random forest - 16S and 18S all taxonomic levels
```{r}
## Format colnames = sampleid, rownames = parameter

# list and read "Actual.tsv" files
list_actual_files = list.files(RF_dir, pattern = "Actual.tsv", full.names = TRUE)
list_actual_dfs = lapply(list_actual_files, read.delim)

# list and read RF "Predictions.tsv" files
list_RF_files = list.files(RF_dir, pattern = "Predictions.tsv", full.names = TRUE)
list_RF_dfs = lapply(list_RF_files, read.delim)

# Function to transpose and format each dataframe and include the file name
convert_table_long <- function(df, file_name) {
  df <- df %>%
    rownames_to_column(var = "parameter") %>%  # Convert row names to a column
    pivot_longer(-parameter, names_to = "sample_id", values_to = "value") %>%
    mutate(source_file = file_path_sans_ext(basename(file_name)))  # Add file name (without extension)
  return(df)
}

# Apply the function to all "Actual.tsv" dataframes, adding file names
actual_dfs <- mapply(convert_table_long, list_actual_dfs, list_actual_files, SIMPLIFY = FALSE)

# Apply the function to all "Predictions.tsv" dataframes, adding file names
RF_dfs <- mapply(convert_table_long, list_RF_dfs, list_RF_files, SIMPLIFY = FALSE)

# Check sample IDs in predictions vs actual match
identical(
  actual_dfs[[1]] %>% select(sample_id) %>% pull(),
  RF_dfs[[1]] %>% select(sample_id) %>% pull()
)

# View an example dataframe
head(actual_dfs[[1]])

# Combine all actual values into one dataframe
actual_combined <- bind_rows(actual_dfs)

actual_combined <- actual_combined %>% 
  mutate(model = str_remove(source_file, "_Actual"))

# Combine all RF predictions into one dataframe
RF_combined <- bind_rows(RF_dfs)

RF_combined <- RF_combined %>% 
  mutate(model = str_remove(source_file, "_Predictions"))

# Merge actual and RF prediction tables by "sample_id" and "variable"
RF_table <- left_join(actual_combined, RF_combined, 
                       by = c("sample_id", "parameter", "model"),
                       suffix = c("_actual", "_predicted"))

# View the combined table
head(RF_table)
```

## TabPNF, amplicon data - 16S and 18S ASVs
```{r}
## Format colnames = sampleid, rownames = parameter
## The models are taken from the file names so TabPNF files were renamed to contain "TabPNF" instead of "RF10Fold"

# list and read "Actual.tsv" files
list_actual_files = list.files(PNF_dir, pattern = "Actual.tsv", full.names = TRUE)
list_actual_dfs = lapply(list_actual_files, read.delim)

# list and read RF "Predictions.tsv" files
list_PNF_files = list.files(PNF_dir, pattern = "Predictions.tsv", full.names = TRUE)
list_PNF_dfs = lapply(list_PNF_files, read.delim)

# Apply the function to all "Actual.tsv" dataframes, adding file names
actual_dfs <- mapply(convert_table_long, list_actual_dfs, list_actual_files, SIMPLIFY = FALSE)

# Apply the function to all "Predictions.tsv" dataframes, adding file names
PNF_dfs <- mapply(convert_table_long, list_PNF_dfs, list_PNF_files, SIMPLIFY = FALSE)

# Check sample IDs in predictions vs actual match
identical(
  actual_dfs[[1]] %>% select(sample_id) %>% pull(),
  PNF_dfs[[1]] %>% select(sample_id) %>% pull()
)

# View an example dataframe
head(actual_dfs[[1]])

# Combine all actual values into one dataframe
actual_combined <- bind_rows(actual_dfs)

actual_combined <- actual_combined %>% 
  mutate(model = str_remove(source_file, "_Actual"))

# Combine all RF predictions into one dataframe
PNF_combined <- bind_rows(PNF_dfs)

PNF_combined <- PNF_combined %>% 
  mutate(model = str_remove(source_file, "_Predictions"))

# Merge actual and RF prediction tables by "sample_id" and "variable"
PNF_table <- left_join(actual_combined, PNF_combined, 
                       by = c("sample_id", "parameter", "model"),
                       suffix = c("_actual", "_predicted"))

# View the combined table
head(PNF_table)
```

## DeepMicro representations, amplicon data - 16S and 18S ASVs
```{r}
## Format colnames = sampleid, rownames = parameter

# list and read "Actual.tsv" files
list_actual_files = list.files(Deep_dir, pattern = "Actual.tsv", full.names = TRUE)
list_actual_dfs = lapply(list_actual_files, read.delim)

# list and read RF "Predictions.tsv" files
list_deep_files = list.files(Deep_dir, pattern = "Predictions.tsv", full.names = TRUE)
list_deep_dfs = lapply(list_deep_files, read.delim)

# Apply the function to all "Actual.tsv" dataframes, adding file names
actual_dfs <- mapply(convert_table_long, list_actual_dfs, list_actual_files, SIMPLIFY = FALSE)

# Apply the function to all "Predictions.tsv" dataframes, adding file names
deep_dfs <- mapply(convert_table_long, list_deep_dfs, list_deep_files, SIMPLIFY = FALSE)

# Check sample IDs in predictions vs actual match
identical(
  actual_dfs[[1]] %>% select(sample_id) %>% pull(),
  deep_dfs[[1]] %>% select(sample_id) %>% pull()
)

# View an example dataframe
head(actual_dfs[[1]])

# Combine all actual values into one dataframe
actual_combined <- bind_rows(actual_dfs)

actual_combined <- actual_combined %>% 
  mutate(model = str_remove(source_file, "_Actual"))

# Combine all RF predictions into one dataframe
deep_combined <- bind_rows(deep_dfs)

deep_combined <- deep_combined %>% 
  mutate(model = str_remove(source_file, "_Predictions"))

# Merge actual and RF prediction tables by "sample_id" and "variable"
deep_table <- left_join(actual_combined, deep_combined, 
                       by = c("sample_id", "parameter", "model"),
                       suffix = c("_actual", "_predicted"))

deep_table <- deep_table %>%
  mutate(model = str_remove(model, "_filtered_norm_seqtab")) %>%
  mutate(model = str_remove(model, "_pivoted_march_2025_rep_RF10fold"))

# View the combined table
head(deep_table)
```

## XGBoost, amplicon data - 16S and (18S ASVs - still running)
```{r}
## Format colnames = sampleid, rownames = parameter

# list and read "Actual.tsv" files
list_actual_files = list.files(XGB_dir, pattern = "Actual.tsv", full.names = TRUE)
list_actual_dfs = lapply(list_actual_files, read.delim)

# list and read RF "Predictions.tsv" files
list_XGB_files = list.files(XGB_dir, pattern = "Predictions.tsv", full.names = TRUE)
list_XGB_dfs = lapply(list_XGB_files, read.delim)

# Apply the function to all "Actual.tsv" dataframes, adding file names
actual_dfs <- mapply(convert_table_long, list_actual_dfs, list_actual_files, SIMPLIFY = FALSE)

# Filter for predicted parameters
filtered_actual_dfs <- lapply(actual_dfs, function(df) {
  df %>% filter(parameter %in% list_of_parameters)
})

# Apply the function to all "Predictions.tsv" dataframes, adding file names
XGB_dfs <- mapply(convert_table_long, list_XGB_dfs, list_XGB_files, SIMPLIFY = FALSE)

# Check sample IDs in predictions vs actual match
identical(
  filtered_actual_dfs[[1]] %>% select(sample_id) %>% pull(),
  XGB_dfs[[1]] %>% select(sample_id) %>% pull()
)

# View an example dataframe
head(filtered_actual_dfs[[1]])

# Combine all actual values into one dataframe
actual_combined <- bind_rows(filtered_actual_dfs)

actual_combined <- actual_combined %>% 
  mutate(model = str_remove(source_file, "_Actual"))

# Combine all RF predictions into one dataframe
XGB_combined <- bind_rows(XGB_dfs)

XGB_combined <- XGB_combined %>% 
  mutate(model = str_remove(source_file, "_Predictions"))

# Merge actual and RF prediction tables by "sample_id" and "variable"
XGB_table <- left_join(actual_combined, XGB_combined, 
                       by = c("sample_id", "parameter", "model"),
                       suffix = c("_actual", "_predicted"))

# View the combined table
head(XGB_table)
```

## All data
```{r}
# Combine all results
all_predictions <- bind_rows(RF_table, PNF_table, deep_table, XGB_table)

# # Remove decimals from predicted year
# all_predictions <- all_predictions %>%
#   filter(parameter == "year") %>%
#   mutate(value_predicted = as.integer(value_predicted))
```


## Calculate Pearson correlation
```{r}
all_predictions_corr <- all_predictions %>%
  group_by(model, parameter) %>%
  summarise(
    # Pearson correlation using cor.test()
    Pearson_test = list(cor.test(value_actual, value_predicted, method = "pearson")),
    Pearson_corr = Pearson_test[[1]]$estimate,  # Extract correlation coefficient
    R2_pearson = Pearson_corr^2,  # Squared correlation (R² from Pearson)
    Pearson_pval = Pearson_test[[1]]$p.value,  # P-value to check significance
    ss_id = sum((value_actual - value_predicted)^2, na.rm = TRUE), # sum of squared errors
    ss_tot = sum((value_actual - mean(value_actual, na.rm = TRUE))^2, na.rm = TRUE), # total sum of squares
    r_squared = ifelse(ss_tot != 0, 1 - (ss_id / ss_tot), 0)
  ) %>%
  ungroup() %>%
  select(-Pearson_test, -ss_id, -ss_tot)

# View the updated dataframe
head(all_predictions_corr)
```

## Krzysztof
```{r}
# R squared
rsquared <- function(predictions, actual) {
  ## Take only complete values from x and y
  i = which(!is.na(predictions) & !is.na(actual))
  predictions = predictions[i]
  actual = actual[i]
  if(length(predictions) == 0 || length(actual) == 0){
    return(NA)
  }
  # Calculate the sum of squares along the identity line
  ss_id <- sum((actual - predictions)^2)
  
  # Calculate the sum of squares of the residuals
  ss_tot <- sum((actual - mean(actual))^2)
  
  # Calculate the correlation coefficient
  if(ss_tot != 0){
    r_squared <- 1 - (ss_id / ss_tot)
    }else if (ss_tot == 0){
    r_squared = 0
    }
  
  return(r_squared)
}

# scatterplot
pdf(paste0(plot_subfolder, parameter, ".pdf"), width = 6, height = 6)
  plot(data$true_values, data$predictions, xlab = 'Data', ylab = 'Predictions', main = parameter,
       bg = alpha('#2C7FB8', 0.5), pch = 21,
       xlim = c(min(data),max(data)), ylim = c(min(data),max(data)))
  lines(c(0,max(data)), c(0,max(data)))
  dev.off()
```

## Plot the data
### Box plot - Random forest, predictions for all parameters
```{r}
all_predictions_corr %>%
  filter(str_detect(model, "RF10fold")) %>%
  ggplot(aes(x = parameter, y = r_squared, fill = parameter)) +
    geom_boxplot() +
    labs(
      x = "Model",
      y = "Pearson_corr"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Add metadata to dataframe from plotting
```{r}
# gene - 16S or 18S
all_predictions_meta <- all_predictions_corr %>%
  mutate(gene = ifelse(str_detect(model, "18S"), "18S", "16S"))

# Add model to new column for plotting - Random_forest, TabPNF or autoencoder
all_predictions_meta <- all_predictions_meta %>%
  mutate(method = case_when(
    str_detect(model, "RF10fold") ~ "Random_forest",
    str_detect(model, "TabPNF") ~ "TabPNF",
    str_detect(model, "^AE") ~ "AE_encoding",
    str_detect(model, "^CAE") ~ "CAE_encoding",
    str_detect(model, "^DAE") ~ "DAE_encoding",
    str_detect(model, "^VAE") ~ "VAE_encoding",
    str_detect(model, "XGBoost") ~ "XGBoost"
  ))

# Add named taxonomic levels
all_predictions_meta <- all_predictions_meta %>%
  mutate(tax_level = case_when(
    str_detect(model, "16S_2") ~ "phylum",
    str_detect(model, "16S_3") ~ "class",
    str_detect(model, "16S_4") ~ "order",
    str_detect(model, "16S_5") ~ "family",
    str_detect(model, "16S_6") ~ "genus",
    str_detect(model, "16S_7") ~ "species",
    str_detect(model, "seqtab_16S") ~ "ASV",
    str_detect(model, "18S_2") ~ "supergroup",
    str_detect(model, "18S_3") ~ "division",
    str_detect(model, "18S_4") ~ "subdivision",
    str_detect(model, "18S_5") ~ "class",
    str_detect(model, "18S_6") ~ "order",
    str_detect(model, "18S_7") ~ "family",
    str_detect(model, "18S_8") ~ "genus",
    str_detect(model, "18S_9") ~ "species",
    str_detect(model, "seqtab_18S") ~ "ASV",
    TRUE ~ "ASV"
  ))

# Double check the columns are correct
all_predictions_meta %>% distinct(method, tax_level)

# Define the order for each amplicon_type
order_16S <- c("phylum", "class", "order", "family", "genus", "species", "ASV")
order_18S <- c("supergroup", "division", "subdivision", "class", "order", "family", "genus", "species", "ASV")

# Apply the order for amplicon_type
all_predictions_meta <- all_predictions_meta %>%
  mutate(tax_level = case_when(
    gene == "16S" ~ factor(tax_level, levels = order_16S),
    gene == "18S" ~ factor(tax_level, levels = order_18S)))

# Create gene + tax level column
all_predictions_meta <- all_predictions_meta %>% unite("type", gene, tax_level, sep = "_", remove = F)

# Define the order of gene + taxonomic level
order_type <- c("16S_phylum", "16S_class", "16S_order", "16S_family", "16S_genus", "16S_species", "16S_ASV", "18S_supergroup", "18S_division", "18S_subdivision", "18S_class", "18S_order", "18S_family", "18S_genus", "18S_species", "18S_ASV")

# Apply the order of gene + taxonomic level
all_predictions_meta <- all_predictions_meta %>%
  mutate(type = factor(type, levels = order_type))

# Apply the order of gene + taxonomic level
# Remove NA values from paramteres not in list (e.g., yday_xcord)
all_predictions_meta <- all_predictions_meta %>%
  mutate(parameter = factor(parameter, levels = list_of_parameters)) %>%
  filter(!(is.na(parameter)))

all_predictions_meta %>% distinct(parameter)

plot_data <- all_predictions_corr %>%
  left_join(
    all_predictions_meta %>%
      select(model, parameter, type, gene, method, tax_level) %>%
      distinct(),
    by = c("model", "parameter")
  )

# Mean R2
plot_data %>% group_by(type, method) %>% 
  summarise(mean_R = mean(r_squared)) %>% 
  arrange(desc(mean_R))


plot_data %>% filter(method ==  "Random_forest") %>% filter(type == "16S_ASV") %>% arrange(desc(r_squared)) %>% summarise(mean(r_squared))

plot_data %>% filter(method ==  "XGBoost") %>% filter(type == "16S_ASV") %>% arrange(desc(r_squared)) %>% summarise(mean(r_squared))
```

### Figure 2AB - Box plot - Random Forest, predictions for all parameters
```{r}
# change so that the colour or shape size is the taxonomic level
Fig2A <- plot_data %>%
  filter(method == "Random_forest") %>%
  filter(parameter %in% physical_parameters) %>%
ggplot(aes(x = gene, y = r_squared, fill = gene)) +
    geom_boxplot(outlier.shape = NA, alpha = 0) +
    geom_point(aes(color = tax_level, shape = tax_level), alpha = 1, size = 2) +
    scale_shape_manual(values = custom_shapes) +
    scale_color_brewer(palette = "Paired") +
    scale_y_continuous(limits = c(0, 1), expand = c(0, 0), breaks = seq(0, 1, by = 0.2)) +
    theme_minimal() +
    #theme_cowplot() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, colour = "black", size = 10),
          axis.text.y = element_text(color = "black", size = 10),
          panel.border = element_rect(colour = "black", fill = NA),
          axis.ticks = element_line(colour = "black"),
          axis.title = element_text(size = 12),
          legend.title = element_text(size = 12, colour = "black"),
          legend.text = element_text(size = 10, colour = "black"),
          strip.background = element_blank(),
          strip.text = element_text(size = 10)) +
    guides(fill = "none",
           shape = guide_legend(ncol = 2)) +
    labs(x = "Model input",
         y = "R2") +
    facet_wrap(~ parameter, nrow = 1)

Fig2B <- all_predictions_meta %>%
  filter(method == "Random_forest") %>%
  filter(parameter %in% nutrient_parameters) %>%
ggplot(aes(x = gene, y = r_squared, fill = gene)) +
    geom_boxplot(outlier.shape = NA, alpha = 0) +
    geom_point(aes(color = tax_level, shape = tax_level), alpha = 1, size = 2) +
    scale_shape_manual(values = custom_shapes) +
    scale_color_brewer(palette = "Paired") +
    scale_y_continuous(limits = c(0, 1), expand = c(0, 0), breaks = seq(0, 1, by = 0.2)) +
    theme_minimal() +
    #theme_cowplot() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, colour = "black", size = 10),
          axis.text = element_text(colour = "black", size = 10),
          panel.border = element_rect(colour = "black", fill = NA),
          axis.ticks = element_line(colour = "black"),
          axis.title = element_text(size = 12),
          legend.title = element_text(size = 12, colour = "black"),
          legend.text = element_text(size = 10, colour = "black"),
          strip.background = element_blank(),
          strip.text = element_text(size = 10)) +
    labs(x = "Model input",
         y = "R2") +
    facet_wrap(~ parameter, nrow = 1)
```

### Box plot - Random Forest at different taxonomic levels (equivalent to Figure 3B)
```{r}
all_predictions_meta %>% 
  filter(method == "Random_forest") %>%
  filter(parameter %in% list_of_parameters) %>%
ggplot(aes(x = type, y = r_squared, fill = tax_level)) +
  geom_boxplot() +
  labs(
    x = "Model",
    y = "R2",
    fill = "Taxonomic level"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, colour = "black", size = 10),
        axis.text.y = element_text(colour = "black", size = 10),
        panel.border = element_rect(colour = "black", fill = NA),
        axis.ticks = element_line(colour = "black"),
        axis.title = element_text(size = 12),
        legend.title = element_text(size = 12, colour = "black"),
        legend.text = element_text(size = 10, colour = "black"))

#ggsave(plot = last_plot(), filename = "Plots/manuscript_draft/Random_forest_taxonomic_levels.png", width = 6, height = 3.5, units = "in")
```
### Figure 2C - Add scatterplots
```{r}
# pdf(paste0(plot_subfolder, parameter, ".pdf"), width = 6, height = 6)
#   plot(data$true_values, data$predictions, xlab = 'Data', ylab = 'Predictions', main = parameter,
#        bg = alpha('#2C7FB8', 0.5), pch = 21,
#        xlim = c(min(data),max(data)), ylim = c(min(data),max(data)))
#   lines(c(0,max(data)), c(0,max(data)))
#   dev.off()

# Salinity  
data <- RF_table %>% 
    filter(str_detect(model, "seqtab_16S")) %>% 
    filter(parameter == "Salinity") %>%
    filter(!(is.na(value_actual)))

min_val <- min(c(data$value_actual, data$value_predicted), na.rm = TRUE)
max_val <- max(c(data$value_actual, data$value_predicted), na.rm = TRUE)

salinity <- ggplot(data, aes(x = value_actual, y = value_predicted)) +
    geom_point(shape = 23, fill = "#D69100",  alpha = 0.7, color = "#D69100", size = 3, stroke = 0.2) +
    geom_abline(slope = 1, intercept = 0) +
    labs(
        x = "Actual",
        y = "Predicted",
        title = "Salinity (R = 0.981)"
    ) +
    coord_equal(xlim = c(min_val, max_val), ylim = c(min_val, max_val)) +
    scale_x_continuous(limits = c(min_val, max_val), expand = c(0, 0)) +
    scale_y_continuous(limits = c(min_val, max_val), expand = c(0, 0)) +
    theme(
    axis.text.x = element_text(angle = 45, hjust = 1, colour = "black", size = 10),
    axis.text.y = element_text(colour = "black", size = 10),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    panel.background = element_blank(),
    axis.ticks = element_line(colour = "black"),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12, colour = "black"),
    legend.text = element_text(size = 10, colour = "black")
  )

# Ammonium
data <- RF_table %>% 
    filter(str_detect(model, "seqtab_16S")) %>% 
    filter(parameter == "NH4") %>%
    filter(!(is.na(value_actual)))

min_val <- min(c(data$value_actual, data$value_predicted), na.rm = TRUE)
max_val <- max(c(data$value_actual, data$value_predicted), na.rm = TRUE)

ammonium <- ggplot(data, aes(x = value_actual, y = value_predicted)) +
    geom_point(shape = 10, color = "#00BDD4", alpha = 1, size = 3, stroke = 0.2) +
    geom_abline(slope = 1, intercept = 0) +
    labs(
        x = "Actual",
        y = "Predicted",
        title = "Ammonium (R = 0.491)"
    ) +
    coord_equal(xlim = c(min_val, max_val), ylim = c(min_val, max_val)) +
    scale_x_continuous(limits = c(min_val, max_val), expand = c(0, 0)) +
    scale_y_continuous(limits = c(min_val, max_val), expand = c(0, 0)) +
    theme(
    axis.text.x = element_text(angle = 45, hjust = 1, colour = "black", size = 10),
    axis.text.y = element_text(colour = "black", size = 10),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    panel.background = element_blank(),
    axis.ticks = element_line(colour = "black"),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12, colour = "black"),
    legend.text = element_text(size = 10, colour = "black")
  )
  
# P_tot
data <- RF_table %>% 
    filter(str_detect(model, "seqtab_16S")) %>% 
    filter(parameter == "P_tot") %>%
    filter(!(is.na(value_actual)))

min_val <- min(c(data$value_actual, data$value_predicted), na.rm = TRUE)
max_val <- max(c(data$value_actual, data$value_predicted), na.rm = TRUE)

p_tot <- ggplot(data, aes(x = value_actual, y = value_predicted)) +
    geom_point(shape = 16, color = "#BC81FF", alpha = 0.7, size = 3, stroke = 0.2) +
    geom_abline(slope = 1, intercept = 0) +
    labs(
        x = "Actual",
        y = "Predicted",
        title = "P_tot (R = 0.826)"
    ) +
    coord_equal(xlim = c(min_val, max_val), ylim = c(min_val, max_val)) +
    scale_x_continuous(limits = c(min_val, max_val), expand = c(0, 0)) +
    scale_y_continuous(limits = c(min_val, max_val), expand = c(0, 0)) +
    theme(
    axis.text.x = element_text(angle = 45, hjust = 1, colour = "black", size = 10),
    axis.text.y = element_text(colour = "black", size = 10),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    panel.background = element_blank(),
    axis.ticks = element_line(colour = "black"),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12, colour = "black"),
    legend.text = element_text(size = 10, colour = "black")
  )

# Alkalinity
data <- RF_table %>% 
    filter(str_detect(model, "seqtab_16S")) %>% 
    filter(parameter == "Alkalinity") %>%
    filter(!(is.na(value_actual)))

min_val <- min(c(data$value_actual, data$value_predicted), na.rm = TRUE)
max_val <- max(c(data$value_actual, data$value_predicted), na.rm = TRUE)

alkalinity <- ggplot(data, aes(x = value_actual, y = value_predicted)) +
    geom_point(shape = 16, color = "#BC81FF", alpha = 0.7, size = 3, stroke = 0.2) +
    geom_abline(slope = 1, intercept = 0) +
    labs(
        x = "Actual value",
        y = "Predicted value",
        title = "Alkalinity (R = 0.979)"
    ) +
    coord_equal(xlim = c(min_val, max_val), ylim = c(min_val, max_val)) +
    theme(
    axis.text.x = element_text(angle = 45, hjust = 1, colour = "black", size = 10),
    axis.text.y = element_text(colour = "black", size = 10),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    panel.background = element_blank(),
    axis.ticks = element_line(colour = "black"),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12, colour = "black"),
    legend.text = element_text(size = 10, colour = "black")
  )

scatter <- cowplot::plot_grid(salinity, p_tot, ammonium, nrow = 1, axis = "bt", align = "hv")

# Get best, worst, and middle
all_predictions_meta %>%
    filter(method == "Random_forest") %>%
    filter(parameter %in% list_of_parameters) %>%
    filter(str_detect(model, "seqtab_16S")) %>% arrange(desc(r_squared))

# Chl
data <- RF_table %>% 
    filter(str_detect(model, "seqtab_16S")) %>% 
    filter(parameter == "Chl") %>%
    filter(!(is.na(value_actual)))

min_val <- min(c(data$value_actual, data$value_predicted), na.rm = TRUE)
max_val <- max(c(data$value_actual, data$value_predicted), na.rm = TRUE)

chl <- ggplot(data, aes(x = value_actual, y = value_predicted)) +
    geom_point(shape = 1, color = "#FF6A9A", alpha = 1, size = 3, stroke = 0.2) +
    geom_abline(slope = 1, intercept = 0) +
    labs(
        x = "Actual",
        y = "Predicted",
        title = "Chl-a (R = 0.249)"
    ) +
    coord_equal(xlim = c(min_val, max_val), ylim = c(min_val, max_val)) +
    theme(
    axis.text.x = element_text(angle = 45, hjust = 1, colour = "black", size = 10),
    axis.text.y = element_text(colour = "black", size = 10),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    panel.background = element_blank(),
    axis.ticks = element_line(colour = "black"),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12, colour = "black"),
    legend.text = element_text(size = 10, colour = "black")
  )

# NO3
data <- RF_table %>% 
    filter(str_detect(model, "seqtab_16S")) %>% 
    filter(parameter == "NO3") %>%
    filter(!(is.na(value_actual)))

min_val <- min(c(data$value_actual, data$value_predicted), na.rm = TRUE)
max_val <- max(c(data$value_actual, data$value_predicted), na.rm = TRUE)

NO3 <- ggplot(data, aes(x = value_actual, y = value_predicted)) +
    geom_point(shape = 13, color = "#00A7FF", alpha = 0.7, size = 3, stroke = 0.2) +
    geom_abline(slope = 1, intercept = 0) +
    labs(
        x = "Actual",
        y = "Predicted",
        title = "NO3 (R = 0.844)"
    ) +
    coord_equal(xlim = c(min_val, max_val), ylim = c(min_val, max_val)) +
    theme(
    axis.text.x = element_text(angle = 45, hjust = 1, colour = "black", size = 10),
    axis.text.y = element_text(colour = "black", size = 10),
    panel.border = element_rect(colour = "black", fill = NA, linewidth = 1),
    panel.background = element_blank(),
    axis.ticks = element_line(colour = "black"),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12, colour = "black"),
    legend.text = element_text(size = 10, colour = "black")
  )

scatter <- cowplot::plot_grid(salinity, NO3, chl, nrow = 1, axis = "bt", align = "hv")
```

### Figure 2C - Random forest at different taxonomic levels - Barplot
```{r}
Fig2C <- all_predictions_meta %>%
  filter(method == "Random_forest") %>%
  filter(parameter %in% list_of_parameters) %>%
  group_by(gene, type) %>%
  summarise(mean_R = mean(r_squared, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = type, y = mean_R, fill = gene)) +
  geom_col(colour = "black") +
  scale_fill_manual(values = c("16S" = "#FC766AFF", "18S" = "#5B84B1FF")) +
  scale_y_continuous(limits = c(0, 0.8), expand = c(0, 0), breaks = seq(0,0.8, by = 0.2)) +
  labs(
    x = "Taxonomic Level",
    y = "Mean (R2)",
    fill = "Gene"
  ) +
  theme_minimal() +
  #theme_cowplot() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, colour = "black", size = 10),
    axis.text.y = element_text(colour = "black", size = 10),
    panel.border = element_rect(colour = "black", fill = NA),
    axis.ticks = element_line(colour = "black"),
    axis.title = element_text(size = 12),
    legend.title = element_text(size = 12, colour = "black"),
    legend.text = element_text(size = 10, colour = "black")
  )
```

### Figure 2 - Combine - v1
```{r}
# Get legends
legend_AB <- get_legend(Fig2A + theme(legend.position = "right"))
legend_C <- get_legend(Fig2C + theme(legend.position = "right"))

# Remove legends from the plots
Fig2A_clean <- Fig2A + theme(legend.position = "none")
Fig2B_clean <- Fig2B + theme(legend.position = "none")
Fig2C_clean <- Fig2C + theme(legend.position = "none")

# Combine legends
combined_legend <- plot_grid(legend_AB, legend_C, ncol = 2)

plot_grid(
    Fig2A,
    Fig2B_clean,
    Fig2C,
    labels = c("A", "B", "C"),
    ncol = 1,
    rel_heights = c(1, 1)
)

ggsave(plot = last_plot(), filename = "Plots/manuscript_draft/Figure2AB_taxonomic_levels_R2.pdf",  height = 8, width = 8)

ggsave(plot = last_plot(), filename = "Plots/manuscript_draft/Figure2AB_taxonomic_levels_R2.png",  height = 8, width = 8)
```

### Figure 3A - Box plot, Random forest vs TabPNF vs XGBoost
```{r}
Fig3a <- all_predictions_meta %>% 
  filter(method == "Random_forest" | method == "TabPNF" | method == "XGBoost") %>%
  filter(tax_level == "ASV") %>%
  filter(parameter %in% list_of_parameters) %>%
ggplot(aes(x = method, y = r_squared)) +
  geom_boxplot(outliers = FALSE) +
  geom_point(aes(color = parameter, shape = parameter), alpha = 1, size = 3) +
  labs(
    x = "",
    y = "R2",
    fill = "Taxonomic level"
  ) +
  scale_shape_manual(values = custom_shapes) +
  scale_y_continuous(breaks = seq(0, 1, by = 0.2), expand = c(0, 0), limits = c(0, 1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, colour = "black", size = 10),
        axis.text.y = element_text(colour = "black", size = 10),
        panel.border = element_rect(colour = "black", fill = NA),
        axis.ticks = element_line(colour = "black"),
        axis.title = element_text(size = 12),
        legend.title = element_text(size = 12, colour = "black"),
        legend.text = element_text(size = 10, colour = "black"),
        legend.position = "bottom") +
  # guides(
  #       fill = guide_legend(ncol = 7),
  #       color = guide_legend(ncol = 7),
  #       shape = guide_legend(ncol = 7)
  #   ) +
  facet_wrap(~ gene, labeller = labeller(gene = gene_labels))

ggsave(plot = Fig3a, filename = "Plots/manuscript_draft/Fig3A_Random_forest_vs_TabPNF_vs_XGBoost_ASVs_R2.pdf", width = 6, height = 5, units = "in")
```

### Statistal significance
```{r}
stat_results <- plot_data  %>% 
  filter(method == "Random_forest" | method == "TabPNF" | method == "XGBoost") %>%
  filter(tax_level == "ASV") %>%
  filter(parameter %in% list_of_parameters) %>%
  group_by(gene) %>%
  wilcox_test(r_squared ~ method)

  # mutate(significance = case_when(
  #   p <= 0.001 ~ "***",
  #   p <= 0.01  ~ "**",
  #   p <= 0.05  ~ "*",
  #   p <= 0.1   ~ ".",
  #   TRUE           ~ "ns"
  # ))

# RF vs XGB - 16S
wide_data <- all_predictions_meta %>%
  filter(method %in% c("Random_forest", "XGBoost")) %>%
  filter(type == "16S_ASV") %>%
  select(parameter, gene, tax_level, method, r_squared, type) %>%
  pivot_wider(names_from = method, values_from = r_squared)

wilcox.test(wide_data$Random_forest, wide_data$XGBoost, paired = TRUE)

# RF vs XGB - 18S
wide_data <- all_predictions_meta %>%
  filter(method %in% c("Random_forest", "XGBoost")) %>%
  filter(type == "18S_ASV") %>%
  select(parameter, gene, tax_level, method, r_squared, type) %>%
  pivot_wider(names_from = method, values_from = r_squared)

wilcox.test(wide_data$Random_forest, wide_data$XGBoost, paired = TRUE)

# TabPNF vs XGB - 16S
wide_data <- all_predictions_meta %>%
  filter(method %in% c("TabPNF", "XGBoost")) %>%
  filter(type == "16S_ASV") %>%
  select(parameter, gene, tax_level, method, r_squared, type) %>%
  pivot_wider(names_from = method, values_from = r_squared)

wilcox.test(wide_data$TabPNF, wide_data$XGBoost, paired = TRUE)

# TabPNF vs XGB - 18S
wide_data <- all_predictions_meta %>%
  filter(method %in% c("TabPNF", "XGBoost")) %>%
  filter(type == "18S_ASV") %>%
  select(parameter, gene, tax_level, method, r_squared, type) %>%
  pivot_wider(names_from = method, values_from = r_squared)

wilcox.test(wide_data$TabPNF, wide_data$XGBoost, paired = TRUE)

# TabPNF vs RF - 16S
wide_data <- all_predictions_meta %>%
  filter(method %in% c("TabPNF", "Random_forest")) %>%
  filter(type == "16S_ASV") %>%
  select(parameter, gene, tax_level, method, r_squared, type) %>%
  pivot_wider(names_from = method, values_from = r_squared)

wilcox.test(wide_data$TabPNF, wide_data$Random_forest, paired = TRUE)

# TabPNF vs RF - 18S
wide_data <- all_predictions_meta %>%
  filter(method %in% c("TabPNF", "Random_forest")) %>%
  filter(type == "18S_ASV") %>%
  select(parameter, gene, tax_level, method, r_squared, type) %>%
  pivot_wider(names_from = method, values_from = r_squared)

wilcox.test(wide_data$TabPNF, wide_data$Random_forest, paired = TRUE)

# RF vs AE - 16S
wide_data <- all_predictions_meta %>%
  filter(method %in% c("Random_forest", "AE_encoding")) %>%
  filter(type == "16S_ASV") %>%
  select(parameter, gene, tax_level, method, r_squared, type) %>%
  group_by(parameter, gene, tax_level, method) %>%
  slice_max(r_squared, n = 1, with_ties = FALSE) %>%
  pivot_wider(names_from = method, values_from = r_squared)

wilcox.test(wide_data$Random_forest, wide_data$AE_encoding, paired = TRUE)

# RF vs CAE - 16S
wide_data <- all_predictions_meta %>%
  filter(method %in% c("Random_forest", "CAE_encoding")) %>%
  filter(type == "16S_ASV") %>%
  select(parameter, gene, tax_level, method, r_squared, type) %>%
  group_by(parameter, gene, tax_level, method) %>%
  slice_max(r_squared, n = 1, with_ties = FALSE) %>%
  pivot_wider(names_from = method, values_from = r_squared)

wilcox.test(wide_data$Random_forest, wide_data$CAE_encoding, paired = TRUE)

# RF vs DAE - 16S
wide_data <- all_predictions_meta %>%
  filter(method %in% c("Random_forest", "DAE_encoding")) %>%
  filter(type == "16S_ASV") %>%
  select(parameter, gene, tax_level, method, r_squared, type) %>%
  group_by(parameter, gene, tax_level, method) %>%
  slice_max(r_squared, n = 1, with_ties = FALSE) %>%
  pivot_wider(names_from = method, values_from = r_squared)

wilcox.test(wide_data$Random_forest, wide_data$DAE_encoding, paired = TRUE)

# RF vs VAE - 16S
wide_data <- all_predictions_meta %>%
  filter(method %in% c("Random_forest", "VAE_encoding")) %>%
  filter(type == "16S_ASV") %>%
  select(parameter, gene, tax_level, method, r_squared, type) %>%
  group_by(parameter, gene, tax_level, method) %>%
  slice_max(r_squared, n = 1, with_ties = FALSE) %>%
  pivot_wider(names_from = method, values_from = r_squared)

wilcox.test(wide_data$Random_forest, wide_data$VAE_encoding, paired = TRUE)

# RF vs AE - 18S
wide_data <- all_predictions_meta %>%
  filter(method %in% c("Random_forest", "AE_encoding")) %>%
  filter(type == "18S_ASV") %>%
  select(parameter, gene, tax_level, method, r_squared, type) %>%
  group_by(parameter, gene, tax_level, method) %>%
  slice_max(r_squared, n = 1, with_ties = FALSE) %>%
  pivot_wider(names_from = method, values_from = r_squared)

wilcox.test(wide_data$Random_forest, wide_data$AE_encoding, paired = TRUE)

# RF vs CAE - 18S
wide_data <- all_predictions_meta %>%
  filter(method %in% c("Random_forest", "CAE_encoding")) %>%
  filter(type == "18S_ASV") %>%
  select(parameter, gene, tax_level, method, r_squared, type) %>%
  group_by(parameter, gene, tax_level, method) %>%
  slice_max(r_squared, n = 1, with_ties = FALSE) %>%
  pivot_wider(names_from = method, values_from = r_squared)

wilcox.test(wide_data$Random_forest, wide_data$CAE_encoding, paired = TRUE)

# RF vs DAE - 18S
wide_data <- all_predictions_meta %>%
  filter(method %in% c("Random_forest", "DAE_encoding")) %>%
  filter(type == "18S_ASV") %>%
  select(parameter, gene, tax_level, method, r_squared, type) %>%
  group_by(parameter, gene, tax_level, method) %>%
  slice_max(r_squared, n = 1, with_ties = FALSE) %>%
  pivot_wider(names_from = method, values_from = r_squared)

wilcox.test(wide_data$Random_forest, wide_data$DAE_encoding, paired = TRUE)

# RF vs VAE - 18S
wide_data <- all_predictions_meta %>%
  filter(method %in% c("Random_forest", "VAE_encoding")) %>%
  filter(type == "18S_ASV") %>%
  select(parameter, gene, tax_level, method, r_squared, type) %>%
  group_by(parameter, gene, tax_level, method) %>%
  slice_max(r_squared, n = 1, with_ties = FALSE) %>%
  pivot_wider(names_from = method, values_from = r_squared)

wilcox.test(wide_data$Random_forest, wide_data$VAE_encoding, paired = TRUE)
```

<!--
```{r}
all_predictions_meta %>% 
    filter(method == "Random_forest" | method == "TabPNF" | method == "XGBoost") %>%
    filter(tax_level == "ASV") %>%
    filter(parameter %in% list_of_parameters) %>%
    ggplot(aes(x = method, y = Pearson_corr)) +
    geom_boxplot(outliers = FALSE) +
    geom_point(aes(color = parameter, shape = parameter), alpha = 0.6, size = 3) +
    labs(
        x = "",
        y = "Pearson correlation (R)",
        fill = "Taxonomic level"
    ) +
    scale_shape_manual(values = custom_shapes) +
    stat_pvalue_manual(
        stat_results_annot,
        label = "label",
        y.position = "y.position",
        tip.length = 0.02,
        bracket.size = 0.5,
        size = 4
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, colour = "black", size = 10),
          axis.text.y = element_text(colour = "black", size = 10),
          panel.border = element_rect(colour = "black", fill = NA),
          axis.ticks = element_line(colour = "black"),
          axis.title = element_text(size = 12),
          legend.title = element_text(size = 12, colour = "black"),
          legend.text = element_text(size = 10, colour = "black"),
          legend.position = "bottom") +
    guides(
        fill = guide_legend(ncol = 7),
        color = guide_legend(ncol = 7),
        shape = guide_legend(ncol = 7)
    ) +
    facet_wrap(~ gene)
```
-->

### Figure 3B - Box plot, Random forest vs encoders
```{r}
fig3b_order <- c("Random_forest", "AE_encoding", "CAE_encoding", "DAE_encoding", "VAE_encoding")

Fig3b <- all_predictions_meta %>% 
    filter(str_detect(method, "Random_forest|encoding")) %>%
    filter(tax_level == "ASV") %>%
    filter(parameter %in% list_of_parameters) %>%
    group_by(method, parameter, gene) %>%
    slice_max(r_squared, n = 1, with_ties = FALSE) %>%
    mutate(method = factor(method, levels = fig4b_order)) %>%
  ggplot(aes(x = method, y = Pearson_corr)) +
    geom_boxplot(outliers = FALSE) +
    geom_point(aes(color = parameter, shape = parameter), alpha = 1, size = 3) +
    labs(
        x = "",
        y = "R2",
        fill = "Taxonomic level"
    ) +
    scale_shape_manual(values = custom_shapes) +
    scale_y_continuous(breaks = seq(0, 1, by = 0.2), expand = c(0, 0), limits = c(0, 1)) +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 45, hjust = 1, colour = "black", size = 10),
        axis.text.y = element_text(colour = "black", size = 10),
        panel.border = element_rect(colour = "black", fill = NA),
        axis.ticks = element_line(colour = "black"),
        axis.title = element_text(size = 12),
        legend.title = element_text(size = 12, colour = "black"),
        legend.text = element_text(size = 10, colour = "black"),
        legend.position = "bottom"
    ) +
    # guides(
    #     fill = guide_legend(ncol = 7),
    #     color = guide_legend(ncol = 7),
    #     shape = guide_legend(ncol = 7)
    # ) +
    facet_wrap(~ gene, labeller = labeller(gene = gene_labels))

ggsave(plot = Fig3b, filename = "Plots/manuscript_draft/Fig3b_encoders_ASVs_R2.png", width = 6, height = 5, units = "in")
```

### Figure 3 - Combine
```{r}
legend <- get_legend(
  Fig3a +
    theme(legend.position = "right",
          plot.margin = margin(100, 100, 100, 100, unit = "pt")))

combined <- plot_grid(
  Fig3a + theme(legend.position = "none"),
  Fig3b + theme(legend.position = "none"),
  labels = c("A", "B"),
  ncol = 2,
  align = 'hv'
)

Fig3 <- plot_grid(
  combined,
  legend,
  ncol = 1,
  rel_heights = c(1, 0.2)
)

ggsave(plot = last_plot(), filename = "Plots/manuscript_draft/Figure3_compareASV_machine_and_encoders_R2.png",  height = 5, width = 8)
```

# Supplementary Figures
## Histogram of input data
```{r}
# All parameters
all_parameters <- c("Longitude", "Latitude", "Salinity", "Temperature", "pH", "Alkalinity", "Secchi_depth", "SiO3", "N_tot", "DIN", "NH4", "NO2", "NO3", "NO3_NO2", "P_tot", "Phosphate", "DOC", "Humus", "Chl")

# Join performance data (R2) to long data by Parameter
responses_df <- responses_matrix %>%
  as.data.frame() %>%
  rownames_to_column(var = "parameter") %>%
  as_tibble() %>%
  pivot_longer(2:ncol(.), names_to = "sample", values_to = "value") %>%
  filter(is.finite(value)) %>%
  filter(parameter %in% all_parameters)

responses_df <- responses_df %>%
  group_by(parameter) %>%
  mutate(parameter_n = paste0(parameter, " (n = ", n(), ")")) %>%
  ungroup()

# Create histograms with R² as subtitle or annotation
ggplot(responses_df, aes(x = value)) +
  geom_histogram(bins = 20, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Physicochemical data",
       x = "value",
       y = "number of samples") +
  theme(strip.text = element_text(size = 10)) +
  facet_wrap(~ parameter_n, scales = "free")

ggsave(plot = last_plot(), "Plots/manuscript_draft/Supplementary_Figure_Input_data.png", height = 8, width = 9, units = "in")
```

# Combine 2 and 3
```{r}
F2A <- Fig2A + theme(legend.position = "none")
F2B <- Fig2B + theme(legend.position = "none")
F2S <- scatter
F2C <- Fig2C + theme(legend.position = "none")
F3A <- Fig3a + theme(legend.position = "none")
F3B <- Fig3b + theme(legend.position = "none")

F2A_legend <- get_legend(Fig2A)
F3A_legend <- get_legend(Fig3a)

#### V1
# Row 1: A (wide)
row1 <- plot_grid(F2B, F2A_legend, ncol = 2, rel_widths = c(1.7, 0.3), labels = c("A", ""))

# Row 2: B and C (normal width)
row2 <- plot_grid(F2A, F2C, ncol = 2, axis = "tb", align = "hv", rel_widths = c(1.2, 0.8), labels = c("B", "C"))

# Row 3: D, E, and legend (now row 4)
row4 <- plot_grid(F3A, F3B, axis = "tb", align = "hv", labels = c("D", "E"))

# Combine all rows
final_plot <- plot_grid(row1, row2, row3, ncol = 1, axis = "l", align = "l", rel_heights = c(0.9, 1.2, 1.2))

ggsave(plot = final_plot, filename = "Plots/manuscript_draft/Combined_fig_2_R2.pdf", width = 180, height = 200, units = "mm", dpi = 300)

#### V2
row1 <- plot_grid(F2B, labels = "A")
row2 <- plot_grid(F2A, NULL, ncol = 2, labels = c("B", ""), rel_widths = c(0.61, 0.39))
plot_grid(row1, row2, nrow = 2)
row3 <- plot_grid(NULL, F2C, nrow = 1, rel_widths = c(2, 1), labels = c("C", "D"))
row4 <- plot_grid(F3A, F3B, nrow = 1, rel_widths = c(0.8, 1.2), axis = "tb", align = "hv", labels = c("E", "F"))

test_plot <- plot_grid(row1, row2, row3, row4, ncol = 1, axis = "l", align = "l", rel_heights = c(1, 1, 1.2, 1.2))

ggsave(plot = test_plot, filename = "Plots/manuscript_draft/Combined_fig_2_R2_scatter.pdf", width = 180, height = 230, units = "mm", dpi = 300)
ggsave(plot = scatter, filename = "Plots/manuscript_draft/scatter.pdf", width = 180, height = 200, units = "mm", dpi = 300)
```
