---
title: "Plot_Results"
author: "Emma"
date: "2024-03-28"
output: html_document
---

# Packages
```{r packages, include=FALSE}
library(tidyverse)
library(purrr)
library(stringr)
library(MLmetrics)
library(scales)
library(ggpubr)
library(cowplot)
library(tools)
```

# Predictions
```{r}
# Random forest
RF_dir <- "/Users/emma/Projects/EnvPredict/Predictions/DifferentTaxonomicLevels"

# TabPNF
PNF_dir <- "/Users/emma/Projects/EnvPredict/Predictions/TabPNF/DifferentTaxonomicLevels"

# XGBoost

```

# Parameters predicted
```{r}
# list parameters to plot
list_of_parameters <- c("Longitude", "Latitude", "Salinity", "Temperature", "pH", "Alkalinity", "Secchi_depth", "SiO3", "N_tot", "DIN", "NH4", "NO3_NO2", "P_tot", "Phosphate", "DOC", "Humus", "Chl")

# Define custom shapes
custom_shapes <- c(16, 17, 18, 19, 15, 8, 3, 4, 7, 9, 10, 12, 13, 14, 20, 21)
```

# Read in data
## Random forest - 16S and 18S all taxonomic levels
```{r}
## Format colnames = sampleid, rownames = parameter

# list and read "Actual.tsv" files
list_actual_files = list.files(RF_dir, pattern = "Actual.tsv", full.names = TRUE)
list_actual_dfs = lapply(list_actual_files, read.delim)

# list and read RF "Predictions.tsv" files
list_RF_files = list.files(RF_dir, pattern = "Predictions.tsv", full.names = TRUE)
list_RF_dfs = lapply(list_RF_files, read.delim)

# Function to transpose and format each dataframe and include the file name
convert_table_long <- function(df, file_name) {
  df <- df %>%
    rownames_to_column(var = "parameter") %>%  # Convert row names to a column
    pivot_longer(-parameter, names_to = "sample_id", values_to = "value") %>%
    mutate(source_file = file_path_sans_ext(basename(file_name)))  # Add file name (without extension)
  return(df)
}

# Apply the function to all "Actual.tsv" dataframes, adding file names
actual_dfs <- mapply(convert_table_long, list_actual_dfs, list_actual_files, SIMPLIFY = FALSE)

# Apply the function to all "Predictions.tsv" dataframes, adding file names
RF_dfs <- mapply(convert_table_long, list_RF_dfs, list_RF_files, SIMPLIFY = FALSE)

# Check sample IDs in predictions vs actual match
identical(
  actual_dfs[[1]] %>% select(sample_id) %>% pull(),
  RF_dfs[[1]] %>% select(sample_id) %>% pull()
)

# View an example dataframe
head(actual_dfs[[1]])

# Combine all actual values into one dataframe
actual_combined <- bind_rows(actual_dfs)

actual_combined <- actual_combined %>% 
  mutate(model = str_remove(source_file, "_Actual"))

# Combine all RF predictions into one dataframe
RF_combined <- bind_rows(RF_dfs)

RF_combined <- RF_combined %>% 
  mutate(model = str_remove(source_file, "_Predictions"))

# Merge actual and RF prediction tables by "sample_id" and "variable"
RF_table <- left_join(actual_combined, RF_combined, 
                       by = c("sample_id", "parameter", "model"),
                       suffix = c("_actual", "_predicted"))

# View the combined table
head(RF_table)
```

## TabPNF, amplicon data - 16S and 18S ASVs
```{r}
## Format colnames = sampleid, rownames = parameter
## The models are taken from the file names so TabPNF files were renamed to contain "TabPNF" instead of "RF10Fold"

# list and read "Actual.tsv" files
list_actual_files = list.files(PNF_dir, pattern = "Actual.tsv", full.names = TRUE)
list_actual_dfs = lapply(list_actual_files, read.delim)

# list and read RF "Predictions.tsv" files
list_PNF_files = list.files(PNF_dir, pattern = "Predictions.tsv", full.names = TRUE)
list_PNF_dfs = lapply(list_PNF_files, read.delim)

# Apply the function to all "Actual.tsv" dataframes, adding file names
actual_dfs <- mapply(convert_table_long, list_actual_dfs, list_actual_files, SIMPLIFY = FALSE)

# Apply the function to all "Predictions.tsv" dataframes, adding file names
PNF_dfs <- mapply(convert_table_long, list_PNF_dfs, list_PNF_files, SIMPLIFY = FALSE)

# Check sample IDs in predictions vs actual match
identical(
  actual_dfs[[1]] %>% select(sample_id) %>% pull(),
  RF_dfs[[1]] %>% select(sample_id) %>% pull()
)

# View an example dataframe
head(actual_dfs[[1]])

# Combine all actual values into one dataframe
actual_combined <- bind_rows(actual_dfs)

actual_combined <- actual_combined %>% 
  mutate(model = str_remove(source_file, "_Actual"))

# Combine all RF predictions into one dataframe
PNF_combined <- bind_rows(PNF_dfs)

PNF_combined <- PNF_combined %>% 
  mutate(model = str_remove(source_file, "_Predictions"))

# Merge actual and RF prediction tables by "sample_id" and "variable"
PNF_table <- left_join(actual_combined, PNF_combined, 
                       by = c("sample_id", "parameter", "model"),
                       suffix = c("_actual", "_predicted"))

# View the combined table
head(PNF_table)
```

## All data
```{r}
all_predictions <- bind_rows(RF_table, PNF_table)
```


## Calculate Pearson correlation
```{r}
all_predictions_corr <- all_predictions %>%
  group_by(model, parameter) %>%
  mutate(
    # Pearson correlation using cor.test()
    Pearson_test = list(cor.test(value_actual, value_predicted, method = "pearson")),
    Pearson_corr = Pearson_test[[1]]$estimate,  # Extract correlation coefficient
    R2_pearson = Pearson_corr^2,  # Squared correlation (R² from Pearson)
    Pearson_pval = Pearson_test[[1]]$p.value,  # P-value to check significance
  ) %>%
  ungroup() %>%
  select(-Pearson_test)  # Remove objects to keep dataframe clean

# View the updated dataframe
head(all_predictions_corr)
```

## Plot the data
### Box plot - Random forest, predictions for all parameters
```{r}
all_predictions_corr %>%
  filter(str_detect(model, "RF10fold")) %>%
  ggplot(aes(x = parameter, y = Pearson_corr, fill = parameter)) +
    geom_boxplot() +
    labs(
      x = "Model",
      y = "Pearson_corr"
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Add metadata to dataframe from plotting
```{r}
# gene - 16S or 18S
all_predictions_meta <- all_predictions_corr %>%
  mutate(gene = ifelse(str_detect(model, "18S"), "18S", "16S"))

# model - Random_forest or TabPNF
all_predictions_meta <- all_predictions_meta %>%
  mutate(method = ifelse(str_detect(model, "RF10fold"), "Random_forest", "TabPNF"))

# Add named taxonomic levels
all_predictions_meta <- all_predictions_meta %>%
  mutate(tax_level = case_when(
    str_detect(model, "16S_2") ~ "phylum",
    str_detect(model, "16S_3") ~ "class",
    str_detect(model, "16S_4") ~ "order",
    str_detect(model, "16S_5") ~ "family",
    str_detect(model, "16S_6") ~ "genus",
    str_detect(model, "16S_7") ~ "species",
    str_detect(model, "seqtab_16S") ~ "ASV",
    str_detect(model, "18S_2") ~ "supergroup",
    str_detect(model, "18S_3") ~ "division",
    str_detect(model, "18S_4") ~ "subdivision",
    str_detect(model, "18S_5") ~ "class",
    str_detect(model, "18S_6") ~ "order",
    str_detect(model, "18S_7") ~ "family",
    str_detect(model, "18S_8") ~ "genus",
    str_detect(model, "18S_9") ~ "species",
    str_detect(model, "seqtab_18S") ~ "ASV",
    TRUE ~ "other"
  ))

# Define the order for each amplicon_type
order_16S <- c("phylum", "class", "order", "family", "genus", "species", "ASV")
order_18S <- c("supergroup", "division", "subdivision", "class", "order", "family", "genus", "species", "ASV")

# Apply the order for amplicon_type
all_predictions_meta <- all_predictions_meta %>%
  mutate(tax_level = case_when(
    gene == "16S" ~ factor(tax_level, levels = order_16S),
    gene == "18S" ~ factor(tax_level, levels = order_18S)))

# Create gene + tax level column
all_predictions_meta <- all_predictions_meta %>% unite("type", gene, tax_level, sep = "_", remove = F)

# Define the order of gene + taxonomic level
order_type <- c("16S_phylum", "16S_class", "16S_order", "16S_family", "16S_genus", "16S_species", "16S_ASV", "18S_supergroup", "18S_division", "18S_subdivision", "18S_class", "18S_order", "18S_family", "18S_genus", "18S_species", "18S_ASV")

# Apply the order of gene + taxonomic level
all_predictions_meta <- all_predictions_meta %>%
  mutate(type = factor(type, levels = order_type))

# Apply the order of gene + taxonomic level
all_predictions_meta <- all_predictions_meta %>%
  mutate(parameter = factor(parameter, levels = list_of_parameters))

all_predictions_meta %>% distinct(parameter)
```

### Box plot - Random Forest at different taxonomic levels
```{r}
all_predictions_meta %>% 
  filter(method == "Random_forest") %>%
ggplot(aes(x = type, y = Pearson_corr, fill = tax_level)) +
  geom_boxplot() +
  labs(
    x = "Model",
    y = "Pearson correlation (R)",
    fill = "Taxonomic level"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, colour = "black", size = 10),
        axis.text.y = element_text(colour = "black", size = 10),
        panel.border = element_rect(colour = "black", fill = NA),
        axis.ticks = element_line(colour = "black"),
        axis.title = element_text(size = 12),
        legend.title = element_text(size = 12, colour = "black"),
        legend.text = element_text(size = 10, colour = "black"))

ggsave(plot = last_plot(), filename = "Plots/manuscript_draft/Random_forest_taxonomic_levels.png", width = 6, height = 3.5, units = "in")
```

### Box plot - Random forest vs TabPNF (ASVs only)
```{r}
all_predictions_meta %>% 
  filter(tax_level == "ASV") %>%
  filter(parameter %in% list_of_parameters) %>%
ggplot(aes(x = model, y = Pearson_corr)) +
  geom_boxplot() +
  geom_point(aes(color = parameter, shape = parameter), alpha = 0.6, size = 3) +
  labs(
    x = "Model",
    y = "Pearson correlation (R)",
    fill = "Taxonomic level"
  ) +
  scale_shape_manual(values = custom_shapes) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, colour = "black", size = 10),
        axis.text.y = element_text(colour = "black", size = 10),
        panel.border = element_rect(colour = "black", fill = NA),
        axis.ticks = element_line(colour = "black"),
        axis.title = element_text(size = 12),
        legend.title = element_text(size = 12, colour = "black"),
        legend.text = element_text(size = 10, colour = "black")) +
  guides(
        fill = guide_legend(ncol = 2),
        color = guide_legend(ncol = 2),
        shape = guide_legend(ncol = 2)
    )

ggsave(plot = last_plot(), filename = "Plots/manuscript_draft/Random_forest_vs_TabPNF_ASVs.png", width = 6, height = 5, units = "in")
```

### Box plot - Random forest - 16S vs 18S, all parameters
```{r}
all_predictions_meta %>% 
  filter(method == "Random_forest") %>%
  filter(parameter %in% list_of_parameters) %>%
ggplot(aes(x = gene, y = Pearson_corr, fill = gene)) +
    geom_boxplot(outlier.shape = NA, alpha = 0) +
    geom_point(aes(color = tax_level, shape = tax_level), alpha = 0.6, size = 3) +
    scale_shape_manual(values = custom_shapes) +
    theme_minimal() +
    theme_cowplot() +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
          strip.background = element_blank(),
          strip.text = element_text(size = 10, margin = margin(t = 1, b = 1))) +
    labs(x = "Taxonomic level",
         y = "Pearson correlation (R)") +
    facet_wrap(~ parameter, nrow = 1)

ggsave(plot = last_plot(), filename = "Plots/manuscript_draft/Random_forest_16S_vs_18S_all_parameters.png",  height = 6, width = 18)
```
